Ribosome subunit attrition and activation of the p53–MDM4 axis dominate the response of MLL-rearranged cancer cells to WDR5 WIN site inhibition

Version 1

Abstract

Summary

The chromatin-associated protein WDR5 is a promising target for cancer drug discovery, with most efforts blocking an arginine-binding cavity on the protein called the “WIN” site that tethers WDR5 to chromatin. WIN site inhibitors (WINi) are active against multiple cancer cell types in vitro, the most notable of which are those derived from MLL-rearranged (MLLr) leukemias. Peptidomimetic WINi were originally proposed to inhibit MLLr cells via dysregulation of genes connected to hematopoetic stem cell expansion. Our discovery and interrogation of small molecule WIN site inhibitors, however, revealed that they act in MLLr cell lines to suppress ribosome protein gene (RPG) transcription, induce nucleolar stress, and activate p53. Because there is no precedent for an anti-cancer strategy that specifically targets RPG expression, we took an integrated multi-omics approach to further interrogate the mechanism of action of WINi in MLLr cancer cells. We show that WINi induce depletion of the stock of ribosomes, accompanied by a broad translational choke, induction of a DNA damage response, and changes in alternative mRNA splicing that inactivate the p53 antagonist MDM4. We also show that WINi are synergistic with agents including venetoclax and BET-bromodomain inhibitors. Together, these studies reinforce the concept that WINi are a novel type of ribosome-directed anti-cancer therapy and provide a resource to support their clinical implementation in MLLr leukemias and other malignancies.

Introduction

WDR5 is a highly-conserved protein that moonlights in a variety of functions in the nucleus. It rose to prominence as a component of epigenetic writer complexes that deposit histone H3 lysine 4 (H3K4) methylation (Guarnaccia & Tansey, 2018), but was subsequently found to act outside these complexes to facilitate the integrity of the mitotic spindle (Ali et al, 2017), bookmark genes for reactivation after mitosis (Oh et al, 2020), and promote transcription of a subset of ribosomal protein genes [(RPGs); (Bryan et al, 2020)] via recruitment of the oncoprotein transcription factor MYC to chromatin (Thomas et al, 2019). WDR5 is also frequently overexpressed in cancer, where its overexpression correlates with aggressive disease and poor clinical outcomes (Guarnaccia & Tansey, 2018). Accordingly, WDR5 is an auspicious target for inhibition in a range of malignancies including MLL-rearranged (MLLr) leukemias (Aho et al, 2019a; Cao et al, 2014), MYC-driven cancers (Aho et al, 2019b), C/EBPα-mutant leukemias (Grebien et al, 2015), p53 gain-of-function cancers (Zhu et al, 2015), neuroblastomas (Bryan et al., 2020), rhabdoid tumors (Florian et al, 2022), and metastatic breast cancers (Cai et al, 2022).

Although WDR5 PROTACs have been described (Li et al, 2022; Yu et al, 2021), safety concerns over destroying a pan-essential protein such as WDR5 (Siladi et al, 2022) means that most drug discovery efforts have focused on small molecule inhibition of key binding sites on the protein. Some initiatives target a hydrophobic cleft on WDR5 known as the “WDR5-binding motif” (WBM) site (Chacon Simon et al, 2020; Macdonald et al, 2019) that contacts MYC (Thomas et al, 2015). But the majority of efforts target the “WDR5-interaction” (WIN) site of WDR5 (Guarnaccia & Tansey, 2018)—an arginine binding cavity that tethers WDR5 to chromatin (Aho et al., 2019a) and makes contact with partner proteins carrying an arginine-containing “WIN” motif (Guarnaccia et al, 2021). Multiple WIN site inhibitors (WINi) have been described (Aho et al., 2019a; Bolshan et al, 2013; Cao et al., 2014; Chen et al, 2021a; Chen et al, 2021b; Grebien et al., 2015; Karatas et al, 2017; Karatas et al, 2013; Li et al, 2016; Senisterra et al, 2013; Tian et al, 2020; Wang et al, 2018), including those that are orally bioavailable and have anti-tumor activity in vivo (Chen et al., 2021b; Teuscher et al, 2023). Given the intense interest in developing WINi for cancer therapy, and the rapid pace of improvement in these molecules, it is likely that WIN site inhibitors will be ready for clinical vetting in the near future.

That said, controversy remains regarding the mechanism of action of WIN site inhibitors, even in the context of MLLr leukemias, where there is strong empirical support for their utility. MLL-rearranged leukemias are defined by translocation of one copy of MLL1—a gene that encodes one of six MLL/SET proteins that are the catalytic subunits of the histone methyltransferase (HMT) complexes responsible for H3K4 methylation (Guarnaccia & Tansey, 2018). The near universal retention of a pristine copy of MLL1 in these cancers led to the idea that MLLr leukemias depend on wild-type MLL1 to support the activity of oncogenic MLL1-fusion oncoproteins (Thiel et al, 2010)—a function in turn that depends on insertion of a low affinity WIN motif within MLL1 into the WIN site of WDR5 (Alicea-Velazquez et al, 2016). Consistent with this notion, early peptidomimetic WINi are active against MLLr leukemia cells in vitro and are reported to suppress levels of H3K4 methylation at canonical MLL1-fusion target genes such as the HOXA loci, causing cellular inhibition through a combination of differentiation and apoptosis (Cao et al., 2014). Subsequently, however, wild-type MLL1 was shown to be dispensable for transformation by MLL-fusion oncoproteins (Chen et al, 2017), and our analysis of picomolar small molecule WINi revealed that they act in MLLr cells without inducing significant changes in the expression of HOXA genes or levels of H3K4 methylation (Aho et al., 2019a). Instead, WIN site inhibitors displace WDR5 from chromatin and directly suppress the transcription of ∼50 genes, the vast majority of which are connected to protein synthesis, including half the cohort of RPGs. We also found that WINi provoke nucleolar stress and induce p53-dependent cell death. Based on our findings, we proposed that WIN site inhibitors kill MLLr cells via depletion of part of the ribosome inventory that induces apoptosis via a ribosome biogenesis stress response.

The concept of ribosome-directed cancer therapies is not new (Laham-Karam et al, 2020; Temaj et al, 2022). Besides mTOR and translational inhibitors, one of the most prevalent strategies in this realm is inhibition of ribosomal RNA (rRNA) production or processing, which is a feature of both existing chemotherapies such as platinum-containing compounds (Bruno et al, 2017), as well as newer targeted RNA polymerase I inhibitors (Drygin et al, 2011; Peltonen et al, 2014). Although these agents exert their anti-cancer effects through multiple mechanisms (Laham-Karam et al., 2020), they are generally thought to disrupt the stoichiometry of RNA and protein components of the ribosome, leading to an excess of ribosomal proteins that inactivate MDM2 to induce p53-dependent cancer cell death. The paradigm we developed for WINi is modeled after that of rRNA inhibitors, although it is important to note that a significant point of divergence from rRNA poisons is that in this model WINi induce p53 not by promoting ribosomal protein accumulation, but by causing a selective imbalance in the ribosome subunit inventory. How such an imbalance could lead to p53 induction, as well as other consequences it may have on MLLr cellular processes, remains unknown.

Fortifying understanding of the mechanism of action of WINi in MLLr cancer cells is key to their clinical implementation. At present, there is no precedent for the mechanism we propose, no understanding of the impact of selective ribosome subunit depletion on translation or other tumor-relevant processes, and no expectations for how resistance to WINi could emerge or how their anti-tumor actions could be made more effective. To ameliorate these deficiencies, we took an integrated multiomics approach, combining transcriptional and translational profiling with genome-wide CRISPR screens to probe WINi action in MLLr cells. Our studies show that although the transcriptional effects of WINi on ribosome subunit expression are confined to those RPGs directly regulated by WDR5, effects at the protein level are not, and WIN site inhibition leads to diminution of the entire stock of cytosolic ribosomes. Ribosome subunit attrition is accompanied by a broad translational choke, induction of a DNA damage response, and activation of p53—driven largely via RPL22-dependent alternative splicing of the p53 antagonist MDM4. We also show that WINi are synergistic with approved and targeted agents including venetoclax and BET-bromodomain inhibitors. Collectively, these findings solidify a novel mechanism of action for WIN site inhibitors in MLLr cells and highlight a path for their optimal clinical implementation.

Results

Impact of WINi on the transcriptome of MLLr cancer cells

Our model for the action of WINi in MLLr leukemia cells is based on analysis of two early generation compounds (Aho et al., 2019a): C3 (Kd = 1.3 nM) and C6 (Kd = 100 pM). Subsequently (Tian et al., 2020), we discovered more potent molecules such as C16 (Kd < 20 pM) that have not been extensively profiled. To ask if improvements in the potency of WINi have resulted in divergent activities, we first compared C6 with C16 (Figure 1A). Both molecules bind the WIN site of WDR5 (Figure 1B and Figure 1—figure supplement 1A), but differ in affinity due to a bicyclic dihydroisoquinolinone core that locks C16 into a favorable binding conformation (Figure 1C). Consistent with its higher affinity, C16 is ∼20 times more potent than C6 in inhibiting MV4;11 (MLL–AF4) and MOLM13 (MLL–AF9) leukemia lines (Figure 1—figure supplement 1B–C): In MV4;11 cells, for example, the GI50 for C6 is 1 µM, compared to 46 nM for C16. These differences in potency are reflected at the level of RPG suppression. Using a target engagement assay (Florian et al., 2022) that measures transcript levels from seven RPGs—five (RPS14, RPS24, RPL26, RPL32, and RPL35) that are always bound by WDR5 and two (RPS11 and RPS14) that are never bound—we observe that maximal suppression of RPG transcripts occurs at ∼2 µM for C6 and ∼100 nM for C16 in MV4;11 (Figure 1D) and MOLM13 (Figure 1—figure supplement 1D) cells. To functionally compare these two inhibitors of different potencies, we used these RPG-normalized doses in all our subsequent studies.

Impact of WINi on the transcriptome of MLLr cancer cells.

(A) Chemical structures of C6 and C16. (B) Crystal structures of C6 or C16 bound to the WIN Site of WDR5 with electrostatic surfaces mapped [PDB IDs: 6E23 (Aho et al., 2019a); 6UCS (Tian et al., 2020)]. The image shows a close-up view of the WIN site. (C) Superimposed WIN site-binding conformations of C6 (green) and C16 (blue). (D) Transcript levels as determined by QuantiGene™ analysis of representative WDR5-bound (color) or non-bound (grayscale) ribosomal protein genes in MV4;11 cells treated with a serial dilution range of either C6 (left) or C16 (right) and relative to DMSO-treated cells (n = 2-3; Mean ± SEM). Vertical dashed line indicates either 2 µM C6 (left) or 100 nM C16 (right). (E) Number of genes with significantly (FDR < 0.05) altered transcript levels following treatment of MV4;11 cells with C6 (2 µM) or C16 (100 nM) for 48 hours, as determined by RNA-Seq (n = 3). See Figure 1—source data 1 for complete output of RNA-Seq analysis. (F) Comparison of gene expression changes elicited by C6 (x-axis) and C16 (y-axis), represented as Log2 fold change (FC) compared to DMSO. WDR5-bound genes are colored red. Locations of RPL22L1 and ZMAT3 are indicated. (G) Overlap of genes with decreased (left) or increased (right) transcript levels in MV4;11 cells treated with C6 or C16. (H) Gene Set Enrichment Analysis (GSEA) showing the distribution of genes suppressed in MV4;11 cells in response to C6 (left) or C16 (right) against the list of all genes bound by WDR5 in those cells (Aho et al., 2019a). NES = normalized enrichment score. (I) Enrichment analysis of genes suppressed (left) or induced (right) by C6 or C16 in MV4;11 cells. KEGG and Hallmark.MSigDB pathways are shown. Fold enrichment of indicated pathways is presented on the x-axis, the number of genes is shown in italics in each bar, and colors represent -Log10 FDR. See Figure 1—source data 2 for additional GSEA (Hallmark) and ORA (Hallmark) analyses of differentially expressed genes. (J) Transcript level changes in WDR5-bound (left) and non-bound (right) RPGs elicited by C6 (top) or C16 (bottom).

We performed RNA-sequencing (RNA-Seq) on MV4;11 cells treated for 48 hours with DMSO, C6, or C16 (Figure 1—figure supplement 1E and Figure 1—source data 1). Both compounds elicit thousands of gene expression changes (Figure 1E), a majority of which are less than two-fold in magnitude (Figure 1—figure supplement 1F). We had previously performed RNA-Seq on MV4;11 cells treated with 2 µM C6 and observed just ∼75 induced and ∼460 reduced transcripts (Aho et al., 2019a). In this earlier work, however, increased variance among replicates made it difficult for as many small gene expression changes to reach statistical significance as in the current study (Figure 1— figure supplement 1G). Comparing the new RNA-Seq datasets, we observe similar effects of C6 and C16 on the MV4;11 cell transcriptome (Figure 1F), with more than 80% of the transcripts altered by C16 altered in the same direction by C6 (Figure 1G). In both cases, suppressed genes are enriched in those bound by WDR5 in MV4;11 cells (Figure 1H); although a majority of transcriptional changes occur at loci bereft of detectable WDR5 binding (Figure 1F). For both compounds, expression of genes connected to protein synthesis, the cell cycle, DNA replication, mTORC signaling, and MYC are reduced, while expression of those connected to chemokine signaling, apoptosis and p53 are induced (Figure 1I and Figure 1—source data 2). Indeed, ∼90 “consensus” p53 target genes (Fischer, 2017) are induced by C6/C16 (Figure 1—figure supplement 2A–B), including the tumor suppressor ZMAT3 [Figure 1F; (Bieging-Rolett et al, 2020)]. Finally, we note that for both compounds, the transcriptional effects on RPG expression are almost entirely confined to those RPGs bound by WDR5 (Figure 1J). The conspicuous exception to this trend is RPL22L1—a paralog of RPL22—mRNA levels which are strongly reduced by C6/C16 (Figure 1F and Figure 1J). This feature of the response is not confined to MLLr cells, as RPL22L1 expression is also decreased by WINi in sensitive rhabdoid tumor cell lines (Florian et al., 2022), but not in the insensitive K562 leukemia line (Bryan et al., 2020) (Figure 1—figure supplement 2C). Despite not being a direct WDR5 target gene, therefore, RPL22L1 expression is recurringly suppressed by WINi in responsive cancer cell lines.

Together, these data reveal that improvements in the potency of WINi have not resulted in substantive changes in their impact on the transcriptome of MLLr cells, and reinforce the concept that inhibition of select RPG expression—and induction of a p53-related transcriptional program—defines the response of the transcriptome to WIN site blockade in this setting.

Impact of WINi on the translatome of MLLr cancer cells

Treatment of MV4;11 cells with 2 µM C6 results in decreased translation, as measured by bulk labeling of nascent polypeptide chains with O-propargyl-puromycin (OPP) (Aho et al., 2019a). To determine more precisely the effects of WIN site inhibition on translational processes, we performed ribosome profiling [Ribo-Seq; (McGlincy & Ingolia, 2017)] in parallel with the RNA-Seq analyses described above. By sequencing ribosome protected fragments (RPF) in treated and control cells, and normalizing to transcript levels from RNA-Seq, we calculated the translation efficiency (TE) of each transcript and used this to determine how C6/C16 influence translation, independent of effects on mRNA abundance.

In these experiments, RPF have the characteristic length of ribosome-protected mRNA fragments (28– 32 nucleotides; Figure 2—figure supplement 1A), are enriched in coding sequences (Figure 2— figure supplement 1B), and map to the expected reading frame Figure 2—figure supplement 1C), all of which indicate successful profiling. In contrast to RNA-Seq, where we see equal numbers of transcript increases and decreases in response to C6/C16, the overwhelming translational effect of these compounds is inhibitory (Figure 2A). Of the ∼10,000 transcripts profiled, between ∼4,500 (C16) and ∼5,900 (C6) transcripts show decreased TE, compared to less than 10 transcripts with increased translation (Figure 2B and Figure 2—source data 1). As we observed in the RNA-Seq, changes in TE are generally less than two-fold (Figure 2A) and there is extensive overlap between the two inhibitors, with ∼90% of the transcripts decreased in TE by C16 also decreased by C6 (Figure 2C). In general, C6/C16 reduce translation of mRNAs in a manner independent of basal translation efficiencies (Figure 2D), although if we bin transcripts according to basal TE we observe that the number of highly-translated transcripts (fourth quartile) impacted by C6/C16 is greater than for those transcripts with lower basal TE (Figure 2E). Within these quartiles, however, the magnitude of reduction in TE is equivalent (Figure 2—figure supplement 1D). Interestingly, mRNAs carrying better matches to the 5’TOP motif—which links translation to mTORC1 signaling (Philippe et al, 2020)—show less decrease in TE compared to those with poorer matches (Figure 2—figure supplement 1E), suggesting that mTORC1-regulated mRNAs may be spared from the full translational effects of WIN site inhibition.

Impact of WINi on the translatome of MLLr cancer cells.

(A) Volcano plots depicting alterations in translation efficiency (TE) induced by 48 hour treatment of MV4;11 cells with either 2 µM C6 (left) or 100 nM C16 (right) compared to DMSO (n = 2; Red indicates FDR < 0.05 and Log2 FC > 0.25), as determined by Ribo-Seq. (B) Number of mRNAs with significantly (FDR < 0.05 and Log2 FC > 0.25) altered TE levels following treatment of MV4;11 cells with C6 (2 µM) or C16 (100 nM) for 48 hours. See Figure 2—source data 1 for complete output of Ribo-Seq analysis. (C) Overlap of mRNAs with significantly decreased TE in response to C6- or C16-treatment. (D) Translation efficiencies (TE) of mRNAs in DMSO-treated MV4;11 cells plotted against translation efficiencies of mRNAs in cells treated with either C6 (left) or C16 (right). Red indicates mRNAs with significantly altered translation efficiencies following inhibitor treatment (FDR < 0.05 and Log2 FC > 0.25). (E) Numbers of differentially-translated mRNAs (ΔTE) in each quartile of genes (stratified by TE in DMSO) in cells treated with C6 (left) or C16 (right). (F) Enrichment analysis of common mRNAs suppressed by C6/C16 at the mRNA (blue) and translational (red; TE) level in MV4;11 cells. Hallmark.MSigDB pathways are shown. The x-axis indicates the number of suppressed genes in each category; the italic numbers are the corresponding FDR. See Figure 2—source data 2 for the full Hallmark.MSigDB analysis, as well as for Reactome and KEGG pathways. (G) Enrichment analysis of mRNAs suppressed translationally by C6/C16 but with no significant changes in mRNA levels. Gene Ontology (GO) Biological Process (BP) and Molecular Function (MF) categories are shown, as well as KEGG pathways. The x-axis displays -Log10 FDR; the number of mRNAs is shown in italics in each bar. See Figure 2—source data 3 for extended enrichment analyses, broken down by TE and mRNA direction changes. (H) TE changes in WDR5-bound (left) and non-bound (right) RPGs elicited by C6 (top) or C16 (bottom).

Interrogating specific changes in TE induced by C6/C16, we see that the biological categories of transcripts with decreased TE echo many of those observed with decreased mRNA levels, but include more genes in each category. For example, manually curating each list for transcripts encoding the ∼60 validated substrates of the protein arginine methyltransferase PRMT5 (Radzisheuskaya et al, 2019) reveals that 42 are translationally suppressed by C6/C16, compared to just 10 that are suppressed at the mRNA level (Figure 2—figure supplement 1F). Probing for Hallmark categories in the Human Molecular Signatures Database [MSigDB; (Liberzon et al, 2015)] uncovers the extent of this phenomenon (Figure 2—source data 2), with categories linked to MYC, E2F, mTORC1 signaling, and the G2M checkpoint (Figure 2F) all represented by more genes in the ribosome profiling than in the RNA-Seq experiments. Within these categories, a majority of genes suppressed by C6/C16 at the mRNA level are also further suppressed translationally (Figure 2—figure supplement 1G). This is not a general trend in the response, however, as fewer than half of the total transcripts with reduced mRNA abundance experience this additional translational inhibition (Figure 2—figure supplement 1H). The finding that TE changes induced by C6/C16 extend the biological characteristics of changes in mRNA abundance may indicate a role for impaired translation in contributing to at least some of the mRNA level differences triggered by WINi. Indeed, comparing transcripts with decreased TE but no mRNA decrease with transcripts with decreased TE and mRNA levels reveals the latter are enriched in so-called “optimal codons” (Wu et al, 2019) that normally promote mRNA stability but are are linked to mRNA instability when translation is inhibited (Figure 2—figure supplement 1I).

Implicit in the previous discussion, a majority of the translational decreases triggered by C6/C16 occur at transcripts for which there are no significant changes in mRNA abundance (Figure 2—figure supplement 1H). Expectedly, genes with decreased TE but no mRNA level changes are enriched in several of the major Hallmark categories described above (Figure 2—source data 3). But we also observe enrichment in genes connected to the proteasome, spliceosome, mRNA surveillance, and translation (Figure 2G). The latter category includes subunits of the mitochondrial ribosome (Figure 2—figure supplement 2A), translation and ribosome biogenesis factors, and an expanded cohort of transcripts from RPGs (Figure 2H). Indeed, compared to mRNA levels, where C6/C16-induced changes are confined (with the exception of RPL22L1) to a decrease in expression of WDR5-bound RPGs, translational effects are not, and some of the most pronounced TE changes occur at non-WDR5 ribosomal protein target genes. Thus, beyond what we have been able to infer from previous studies, WINi causes a widespread reduction in the ability of MLLr cells to efficiently translate mRNAs connected to almost every aspect of protein synthesis and homeostasis.

Impact of WINi on the ribosome inventory of MLLr cancer cells

Based on the finding that WDR5 controls expression of half the RPGs, we speculated that WINi induce a ribosome subunit imbalance that leads to induction of p53 (Aho et al., 2019a). It is also possible, however, that quality control mechanisms deplete the entire inventory of ribosomal proteins during prolonged WIN site blockade. To distinguish between these possibilities, we tracked changes in ribosomal protein levels at two timepoints: 24 hours, when there is no overt cellular response to WIN site inhibition, and 72 hours, when cell proliferation begins to be inhibited (Aho et al., 2019a). The abundance of ribosomal proteins allows for the use of label-free quantitative mass spectrometry [LFQMS; (Cox et al, 2014)] in whole cell lysates to feasibly track ribosome protein levels, while also providing insight into other changes in protein levels promoted by WIN site inhibitor.

In this analysis, we tracked ∼3,200 proteins at each timepoint (Figure 3A and Figure 3—source data 1), ∼850 of which are significantly altered by C16 treatment. Consistent with the subtle effects of WINi on mRNA abundance and TE, most differences in protein levels triggered by C16 are less than two-fold in magnitude (Figure 3B and 3C). At 24 hours, ∼90% of the proteins that change in response to C16 score as increased, whereas by 72 hours this number drops to ∼60% (Figure 3A). Most instances of increased protein levels are transient—fewer than one third of these proteins are still induced at day three (Figure 3D)—whereas a majority of the proteins decreased at day one are also decreased at day three. Enrichment analysis (Figure 3—figure supplement 1A and Figure 3—source data 2) reveals that proteins induced at 24 hours are modestly enriched in those connected to exocytosis and leukocyte activation, as well as mTORC1 signaling and MYC. By 72 hours, we see induction of proteins linked to glycolysis and fatty acid metabolism, as well as apoptosis. Additionally, manual curation reveals that the number of induced p53 target proteins increases over time: 13 are induced at 24 hours, compared to 24 at the three day point (Figure 3E). Commensurate with the onset of a functional response to WINi, therefore, is a modest expansion in the apparent impact of p53 on the proteome, as well as the emergence of apoptotic response indicators.

Impact of WINi on the ribosome inventory of MLLr cancer cells.

(A) Lysates from MV4;11 cells treated 24 or 72 hours with either 0.1% DMSO or 250 nM C16 were subjected to liquid chromatography coupled with tandem mass spectrometry and analyzed by label-free quantification (LFQMS). The table shows the number of proteins detected in DMSO and C16 samples and those with significantly altered levels at each time point (n = 4; adj. p-value < 0.05). See Figure 3—source data 1 for complete output of LFQMS analysis. (B) Volcano plot, showing protein level alterations in cells treated with C16 for 24 hours (red indicates adj. p-value < 0.05). The location of RPL22L1 is indicated. (C) As in (B) but for 72 hour treatment with C16. (D) Overlap of proteins significantly increased (top) or decreased (bottom) following 24- or 72-hour C16-treatment. (E) Protein level alterations induced by C16 in consensus p53 target proteins (Fischer, 2017) at the 24 and 72 hour treatment timepoints. Those proteins only altered in abundance at 24 hours are represented as blue dots; proteins only altered at 72 hours are red; proteins altered at both timepoints are grey. (F) As in (E) but for ribosomal proteins. (G) Changes in expression of proteins encoded by WDR5-bound (left) and non-bound (right) RPGs elicited by 24 (top) or 72 (bottom) hour treatment with C16. Note that, due to the magnitude of change, Log2(FC) for RPL22L1 is presented on a separate scale.

Not surprisingly, proteins that are reduced in abundance at 24 hours are significantly enriched in those linked to the ribosome (Figure 3—figure supplement 1B). This enrichment becomes stronger at 72 hours. We also observe, at 72 hours, suppression of proteins linked to MYC and E2F targets, as well as mTORC1 signaling. In terms of ribosome components, this analysis reveals a progressive decline in the ribosomal protein inventory. Going from 24 to 72 hours, there is an increase in the number of impacted ribosomal subunits as well as in the magnitude of their suppression (Figure 3F), and eventually almost all ribosomal subunits are in deficit, regardless of whether or not they are encoded by a WDR5-bound gene (Figure 3G). Interestingly, RPL22L1 is the most strongly suppressed protein at 72 hours (Figure 3C), with its levels reduced by an order of magnitude after three days of C16 treatment. In sum, this analysis reveals that changes in ribosomal protein levels predicted from our transcriptomic studies manifest in reduced expression of ribosome components. Contrary to our earlier idea that WINi promote ribosome subunit imbalance, however, these data support a simpler model in which these inhibitors ultimately induce attrition of the majority of ribosomal proteins.

A loss of function screen for modulators of the response to WINi

Next, we conducted a loss of function screen to identify genes that modulate the response of MLLr cells to WINi (Figure 4A). First, we carried out a screen using the GeCKOv.2 sgRNA library (Joung et al, 2017), which targets ∼19,000 genes with six sgRNAs each, as well as ∼1,200 miRNAs (four sgRNAs each). After transducing the library into MV4;11 cells expressing Cas9, we treated for two weeks with 2 µM C6, during which time rapidly growing cells emerged within the transduced population (Figure 4— figure supplement 1A). We harvested genomic DNA, performed next-generation sequencing, and compared sgRNA representation before and after C6 treatment. We inventoried genes with significant enrichment/depletion in corresponding sgRNAs in the treated population, removed pan-essential genes (Tsherniak et al, 2017), and created a custom library in which protein-coding “hits” are targeted by four different sgRNAs (Doench et al, 2016). The smaller library was screened against C6 or C16 in Cas9-expressing MV4;11 cells. This two-tiered approach allowed us to efficiently screen two different WIN site inhibitors and to identify high-confidence hits that are validated with up to 10 distinct sgRNAs.

A loss of function screen for modulators of the response to WINi.

(A) Two-tier screen design. In the first tier, Cas9-expressing MV4;11 cells were transduced with a genome-wide sgRNA library and treated with 2 µM C6 until a resistant cell population emerged. sgRNA representation in the pre-treatment population was compared to the post-treatment population. In the second tier, cells were transduced with a custom library of distinct sgRNAs targeting non-pan-essential “hits” from the first tier, cultured in the presence of DMSO, C6, or C16, and sgRNA representation in C6/C16-treated cultures compared to that from DMSO-treated cultures. Created with BioRender.com. (B) Volcano plot, showing gene-level changes in sgRNA representation from the first tier (orange indicates FDR < 0.05). Datapoints corresponding to TP53, RPL22, and CDKN2A are indicated. See Figure 4—source data 1 for full output of the Tier 1 screen. (C) Comparison of gene-level changes in sgRNA representation in C6- and C16-treated populations in the second tier screen, each compared to DMSO-treated populations (red indicates FDR < 0.05; black indicates non-targeting control sgRNAs). See Figure 4— source data 2 for full output of the Tier 2 screen. (D) Top: Overlap of genes from the Tier 2 screen with enriched (left) or depleted (right) sgRNAs in C6- and C16-treated MV4;11 populations, compared to the DMSO control. Bottom: Overlap of genes with enriched (left) or depleted (right) sgRNAs in the first versus second tiers of the screen. “Tier 1” contains only those genes targeted in the Tier 2 screen. “Tier 2” contains the intersection of genes with altered sgRNAs in both the C6 and C16 treatments. (E) Ranked heatmap, representing the mean gene-level Log2 fold change (FC) of sgRNAs from the C6 and C16 treatments in the Tier 2 screen, as well as gene enrichment analysis outputs. Note that “Signal transduction by p53 class mediator” is a GO:BP term (orange); “p53” assignments (yellow) were added by manual curation.

Although the first tier did not discriminate between genes that modulate fitness and those that modulate WINi response, several interesting observations emerged. Guide RNAs corresponding to ∼70 genes were enriched and ∼675 were depleted (Figure 4B), most of the latter of which are pan-essential (Figure 4—source data 1). Satisfyingly, TP53 is the most highly enriched gene in the screen (Figure 4B and Figure 4—figure supplement 1B). CDKN2A also scored as highly enriched in the initial screen (Figure 4B)—specifically those sgRNAs targeting p14ARF(Figure 4—figure supplement 1C– 1D), an inhibitor of the p53 ubiquitin ligase MDM2 (Sherr, 2001). Further support for the importance of p53 is evidenced by network analysis (Chang & Xia, 2023) of the 27 miRNAs flagged as enriched (Figure 4—source data 1), which display connections to p53 (Figure 4—figure supplement 1E). Finally, we note that the second most highly-enriched gene in the first tier encodes a ribosomal protein: RPL22 (Figure 4B and Figure 4—figure supplement 1F). Because of the strong enrichment of sgRNAs against RPL22 and TP53, we removed both genes from the second tier screen.

The second tier screen (Figure 4—figure supplement 1G and Figure 4—source data 2) revealed that the response of MV4;11 cells to C6 and C16 is very similar, both in terms of the enriched/depleted genes and their rankings (Figure 4C). A majority of genes that modulate the response to C6 similarly modulate the response to C16 (Figure 4D). For most of the depleted genes that appear specific to one WINi, similar depletion is observed with the other WIN site inhibitor, but is generally just over the FDR cutoff (Figure 4—figure supplement 1H). But for C6-specific enriched genes we see that most have high FDR values in the C16 samples, arguing that the earlier generation compound has expanded, off-target, activities. Gene ontology (GO) enrichment analysis of the 57 common genes emerging from the screen revealed enrichment in four overlapping categories connected to p53 signaling, apoptosis, the DNA damage response, and histone modifications (Figure 4E and Figure 4—figure supplement 1I). The representation of genes connected to p53 and to apoptosis reinforces the importance of p53-mediated cell death to the response of MLLr cells to WINi. We observe, for example, that loss of function of the p53 antagonist and ubiquitin ligase MUL1 (Jung et al, 2011), increases sensitivity to C6/ C16, whereas loss of canonical p53 effectors NOXA, BAX, and ZMAT3 is associated with a decrease in response. The DNA damage response (DDR) category overlaps with that of p53 but is nonetheless distinct and includes genes encoding the ATM and CHK2 kinases (Blackford & Jackson, 2017) and the FOXM1 transcription factor that activates DDR gene expression networks (Zona et al, 2014). This category also includes two depleted genes, encoding FBXL5—which antagonizes ATM signaling (Chen et al, 2014)—and DYRK1A—a kinase involved in the DDR (Laham et al, 2021), DREAM complex activation (Litovchick et al, 2011), and RPG transcription (Di Vona et al, 2015). Consistent with representation of DDR genes in hits from the screen, C16 induces the DNA damage marker ɣ-H2AX in MV4;11 cells (Figure 4—figure supplement 1J and Figure 4—source data 3), implying that DNA damage, and the ability to mount a DDR, are part of the mechanism of action of WINi in this setting.

A majority of genes in the histone modification category, when disrupted, blunt the response to both WINi (Figure 4E). These genes include those encoding the H3K27 demethylase KMD6A (Lan et al, 2007), the MLL/SET protein KMT2D [MLL2; (Shinsky et al, 2015)], and ARID4A—a component of the mSin3/HDAC1 co-repressor complex (Lai et al, 2001). The most conspicuous sensitizing gene in this group is BRD3, a member of the BET family of proteins that includes BRD2 and BRD4 (Eischer et al, 2023). Interestingly, although BRD4 was not included in the second screen tier as it is pan essential, BRD2 was not included because it was not significantly enriched/depleted in the first tier (Figure 4— source data 1), revealing that the actions of BRD3 in modulating response to WINi are not shared with all family members. Further supporting the importance of BRD3 to the response, we note that SPOP, which targets BET family proteins for proteasomal destruction (Janouskova et al, 2017), is one of the most significantly enriched hits from the screen (Figure 4E).

Collectively, these findings demonstrate functional involvement of the ribosomal protein RPL22 in the response to WINi, confirm the importance of the p53 network to robust inhibition of MLLr cell growth by these agents, and reveal a role for DNA damage in the response to WIN site inhibitor. These findings also identify a number of candidate predictive biomarkers of response that can be further interrogated for their value as patient-selection criteria.

Identification of agents that synergize with WINi in MLLr cells

Given the ways in which resistance to WINi can arise (Figure 4), the most efficacious application of these agents will likely be in combination with other therapies. We therefore asked whether C16 synergizes with 11 approved or targeted agents. Several of the agents were chosen based on the results of our screen. Harmine—an inhibitor of the DYRK1A kinase (Gockler et al, 2009)—and the BET family inhibitor mivebresib (McDaniel et al, 2017) each target the product of sensitizing genes, whereas venetoclax inhibits BCL-2 (Souers et al, 2013)—an inhibitor of BAX, which scored as a resistance gene. We also tested agents connected to the DDR [etoposide, olaparib, and the ATR inhibitor VE821; (Charrier et al, 2011)], protein synthesis and homeostasis (alvespimycin and rapamycin), and p53 (nutlin-3a). Due the the enrichment of PRMT5 substrates in our translational profiling (Figure 2—figure supplement 1F), we queried the PRMT5 inhibitor pemrametostat (Chan-Penebre et al, 2015). And because DOT1L inhibitors suppress not only classic MLL fusion target genes (Bernt & Armstrong, 2011; Daigle et al, 2011) but also RPGs (Lenard et al, 2020), we tested for synergy with the DOT1L inhibitor pinometostat (Daigle et al., 2011). We treated MV4;11 cells with a dose matrix spanning 49 unique dose combinations and quantified synergy δ-scores using the Zero Interaction Potency (ZIP) model (Yadav et al, 2015) (Figure 5A-B, Figure 5—figure supplement 1A and 1B, and Figure 5—source data 1).

Identification of agents that synergize with WINi in MLLr cells.

(A) Peak synergy (> 0) and antagonism (< 0) ZIP Delta (δ) scores from synergy assays in which MV4;11 cells were treated for three days with 49 unique dose combinations of C16 and the indicated compound of interest (n = 4). See Figure 5—source data 1 for numerical ZIP Delta analysis output. (B) Heatmaps of MV4;11 cell growth inhibition at each dose of C16 and the indicated six compounds. The remaining five combinations tested are shown in Figure 5—figure supplement 1. (C) As in (A) but for MOLM13 cells. See Figure 5—source data 1 for numerical ZIP Delta analysis output. (D) As in (B) but for MOLM13 cells. The remaining five combinations tested are shown in Figure 5—figure supplement 2. (E) Number of genes with significantly (FDR < 0.05) altered transcript levels following treatment of MV4;11 cells with C16 (100 nM), mivebresib (Mibv; 2.5 nM), or the combination for 48 hours, as determined by RNA-Seq (n = 3). See Figure 5—source data 2 for complete output of RNA-Seq analysis. (F) UpSet plot, showing the overlap of genes suppressed (left) or induced (right) in response to C16, mivebresib, or the combination. (G) UpSet plot, showing the breakdown of Reactome “Translation” pathway genes suppressed in response to C16, mivebresib, or the combination. (H) Enrichment of Reactome Pathways in genes with increased transcripts following treatment of MV4;11 cells with C16, mivebresib, or the combination. See Figure 5—source data 3 for complete output of enrichment analyses.

In MV4;11 cells, we observe synergy with mivebresib, pemrametostat, pinometostat, etoposide, harmine, and venetoclax. Within this group are three agents selected based on sensitizing targets from the CRISPR screen, providing additional support for the role of DRYK1A, BCL-2/BAX, and BRD3 in the responsiveness to WINi. Of the three agents connected to the DDR, only etoposide displays significant synergy. Agents that target protein synthesis and homeostasis yield mixed results—we observe potent antagonism with the mTOR inhibitor (Raught et al, 2001) rapamycin (peak δ-score −16), while the HSP90 inhibitor alvespimycin (Schnur et al, 1995) is either antagonistic or synergistic, depending on dose (Figure 5A). Finally, we note that—of the agents displaying synergy—four are particularly strong (peak δ-scores >10) and observed at agent doses consistent with on-target activity (Figure 5—figure supplement 1B), suggesting that mivebresib, pemrametostat, pinometostat, and venetoclax should be prioritized for in vivo testing. Focusing on these agents is further justified by our finding that all four are synergistic with C16 in MOLM13 cells (Figure 5C–D and Figure 5—figure supplement 2A and 2B).

To understand how combination with another agent impacts the response to WINi, we profiled MV4;11 cells treated for forty-eight hours with C16 and mivebresib, either as single agents or in combination, at concentrations that yield peak synergy between them (100 nM C16 and 2.5 nM mivebresib). By RNA-Seq (Figure 5—source data 2), it is clear that the functional synergy between C16 and mivebresib is apparent at the transcript level, with more than 6,200 gene expression changes in the combination treatment, compared to less than 1,800 for C16 and 2,700 for mivebresib (Figure 5E). Notable are the very distinct transcriptional profiles induced by each agent alone, with fewer than 200 shared gene expression changes in each direction (Figure 5F). The impact on RPG expression of both agents is additive (Figure 5—figure supplement 3A), but in general we find that the combination of C16 and mivebresib dysregulates similar categories of genes for each agent alone, but with substantially more genes in each category (Figure 5—source data 3). This is clear for genes linked to translation (Figure 5G), to p53 (Figure 5—figure supplement 3B), and to the induction of apoptosis (Figure 5H). Thus, although further investigation is needed, this analysis is consistent with the idea that synergy between C16 and mivebresib results from alterations in the expression of distinct but complementary sets of genes that ultimately conspire to augment induction of p53.

WINi inactivate MDM4 in an RPL22-dependent manner

Despite the importance of p53 in the response to WINi, WIN site inhibitors cause only a slight increase, if any, in p53 levels [Figure 6—figure supplement 1A, Figure 6—source data 1 and (Aho et al., 2019a)]. Interestingly, inactivation or loss of RPL22 in cancer is associated with increased expression of RPL22L1 and inclusion of exon 6 in MDM4 (Ghandi et al, 2019), an event that promotes MDM4 expression by preventing formation of a “short” MDM4 mRNA isoform (MDM4s) that is destroyed by nonsense-mediated decay (Rallapalli et al, 1999). MDM4 is intriguing because it can suppress p53 without altering its stability (Francoz et al, 2006). It is also intriguing because skipping of exon 6 in the MDM4 mRNA is stimulated by ZMAT3 (Bieging-Rolett et al., 2020; Muys et al, 2021) and antagonized by RPL22L1 (Larionova et al, 2022)—two genes that are oppositely regulated by WINi. We therefore asked if WINi induce changes in the levels of mRNA splice isoforms and if this includes MDM4.

WINi inactivate MDM4 in an RPL22-dependent manner.

(A) Differential alternative splicing events affected by C6/C16 treatment of MV4;11 cells were quantified by rMATS. The types of alternative splicing events are cartooned at left, and the number of significantly different events (> 5% Δψ; FDR < 0.05) common to C6/C16 depicted in the graph. See Figure 6—source data 2 for output of rMATS analysis. (B) Sashimi plot quantifying read junctions that span exons 5–7 of MDM4 in MV4;11 cells treated with DMSO (green) or C16 (blue). Numbers in the arcs display junction depth. The location of exons 5, 6, and 7 is depicted at the bottom; skipped exon 6 is highlighted in orange. (C) Viabilities of control (non-targeting: NT) and RPL22 knock out (KO) MV4;11, MOLM13, and K562 cells treated with a serial dilution range of C16 for 72 hours, relative to viability of DMSO-treated cells (n = 3; Mean ± SEM). (D) Western blot analysis of p53 levels in control (NT) and RPL22 knockout (KO) MV4;11 and MOLM13 cells treated with either 0.1% DMSO or C16 (MV4;11, 200 nM; MOLM13, 400 nM) for 72 hours. α-Actinin is loading control. Representative images from three biological replicates shown. Raw unprocessed gel images are presented in Figure 6—source data 5. (E) Heatmap, showing significant changes in the expression of consensus p53 target genes (Fischer, 2017) between the indicated pairwise comparisons of RNA-Seq datasets. Note that only consensus p53 target genes altered in expression by C16 in control (NT) cells are represented. (F) Sashimi plot quantifying read junctions that span exons 5–7 of MDM4 in RPL22KO MV4;11 cells treated with DMSO or C16. Numbers in the arcs display junction depth. The location of exons 5, 6, and 7 is depicted at the bottom; skipped exon 6 is highlighted in orange. Corresponding NT images are presented alongside RPL22KO images in Figure 6—figure supplement 3B). (G) Western blots, comparing the effects of 72 hours of DMSO (DM) or C16 treatment (MV4;11, 200 nM; MOLM13, 400 nM) of control (NT) or RPL22 knockout (KO) MV4;11 (left) or MOLM13 (right) cells on levels of MDM4, p21, RPL22L1, RPL22, and GAPDH (loading control). Representative images from three biological replicates are shown. Raw unprocessed gel images are presented in Figure 6—source data 9.

RNA-Seq data (Figure 1) were interrogated for alternative splicing events (Shen et al, 2014). At an FDR < 0.05 and a threshold of ≥ 5% change in exon inclusion (Δψ), C6 and C16 each result in changes in ∼1,000 differentially-spliced mRNAs (Figure 6—figure supplement 1B), ∼250 of which are shared between the two inhibitors (Figure 6A). Many of these changes reflect events with low read counts or at minor splice sites (Figure 6—source data 2), representative examples of which are presented in Figure 6–figure supplement 1C–D. That said, WINi clearly promote accumulation of MDM4 transcripts in which exon 6 is skipped (Figure 6B). We also observe splicing changes at RPL22L1 itself (Figure 6–figure supplement 1E), where WINi leads to the depletion of transcripts in which exon 2 is spliced to a distal 3’ acceptor site in exon 3. This splicing event encodes the RPL22L1a isoform that is incorporated into ribosomes (Larionova et al., 2022). Splicing to the proximal 3’ acceptor site, which generates a non-ribosomal RPL22L1b isoform that modulates splicing, is insensitive to WINi. We confirmed the impact of C16 on MDM4 and RPL22L1 splice isoforms by semi-quantitative RT-PCR (Figure 6—figure supplement 1F and Figure 6—source data 3) and quantitative RT-PCR (Figure 6—figure supplement 1G). Based on these observations, we conclude that treatment of MLLr cells with WINi promotes the selective loss of transcripts encoding RPL22L1a and MDM4.

The association of RPL22 loss with increased expression of RPL22L1 and inclusion of exon 6 in MDM4 (Ghandi et al., 2019) prompted us to ask how RPL22 contributes to the response of MLLr cells to WINi. Knockout (KO) of RPL22 (Figure 6—figure supplement 2A and Figure 6—source data 4) decreases the sensitivity of MV4;11 and MOLM13 cells to C16 by three- to five-fold compared to non-targeted (NT) control cells (Figure 6C and Figure 6—figure supplement 2B), as well as attenuating the modest induction of p53 protein observed in the MV4;11 line (Figure 6D and Figure 6—source data 5). The response of relatively insensitive (p53-null) K562 cells, in contrast, is unaffected by RPL22 disruption (Figure 6C). RNA-Seq analysis (Figure 6—figure supplement 2C–D and Figure 6—source data 6) reveals that disruption of RPL22 does not impact the effect of WINi on WDR5-bound RPGs (Figure 6— figure supplement 2E), but it does block the effects of C16 on expression of RPL22L1 and ZMAT3 (Figure 6—figure supplement 2D), as well as tempering its ability to suppress genes connected to the cell cycle, mTORC1 signaling, and MYC (Figure 6—figure supplement 2F and Figure 6—source data 7). Notably, RPL22 loss also impairs induction of genes involved in p53 signaling (Figure 6E and Figure 6—figure supplement 2G). We also observe that mitochondrial RPGs are induced by WINi uniquely in RPL22-null cells (Figure 6—figure supplement 2H). We conclude that RPL22 is needed for a majority of the characteristic responses of MLLr cells to WINi, including activation of p53.

Finally, we asked if RPL22 knockout alters patterns of alternative splicing induced by WINi. Thousands of differences were detected in splice isoforms between the various pairwise comparisons (Figure 6— source data 8). In general, RPL22KO cells show fewer C16-induced changes in alternative splicing patterns than NT cells (Figure 6—figure supplement 3A). As we observed above, a majority of the changes reflect events with low read counts or at minor splice sites, with two notable exceptions: MDM4 and RPL22L1. In the absence of WINi, disruption of RPL22 promotes exon 6 retention in MDM4 (Figure 6—figure supplement 3B) and induces expression of the mRNA splice isoform encoding RPL22La (Figure 6—figure supplement 3C). In the presence of C16, disruption of RPL22 mitigates skipping of exon 6 in MDM4 (Figure 6F) and largely blocks suppression of the RPL22L1a-encoding mRNA isoform (Figure 6—figure supplement 3C). Importantly, these changes in mRNA isoforms manifest at the protein level (Figure 6G and Figure 6—source data 9) as we observe that both MDM4 and RPL22L1 protein expression is suppressed by C16, that RPL22L1 is induced by RPL22 disruption, and that loss of RPL22 blocks a majority of the ability of C16 to reduce expression of the MDM4 and RPL22L1 proteins. We also note that loss of RPL22 blocks induction of the p53 target, p21. Taken together, these data demonstrate that RPL22 loss can cause the induction of RPL22L1 and inclusion of exon 6 in MDM4 observed in cancer (Ghandi et al., 2019), and reveal that splicing-driven suppression of MDM4 is a major mechanism through which WINi activate p53 in MLLr cells.

Discussion

Here, we describe an integrated multiomic approach to characterize the mechanism of action of WDR5 WIN site inhibitors in MLL-rearranged cancer cells. By combining transcriptional, translational, and proteomic profiling with genome-wide loss of function screens, we demonstrate the broad impact of WINi on the ribosomal protein complement and translational capacity of MLLr cells, reveal the importance of multiple arms of the p53 response pathway in cellular inhibition, and uncover a role for alternative splicing of MDM4 in activating p53 in this setting. Collectively, these data cast WIN site inhibitors as a novel ribosome-directed anti-cancer therapy and provide insight into patient selection criteria, mechanisms of resistance, and strategies to improve WINi efficacy in the clinic.

The primary targets of WINi, we propose, are the cohort of ∼40 RPGs that are bound by WDR5 in all cell types examined (Aho et al., 2019a; Bryan et al., 2020; Florian et al., 2022). These RPGs are rapidly transcriptionally suppressed in response to WIN site blockade (Aho et al., 2019a; Bryan et al., 2020; Florian et al., 2022) or WDR5 degradation (Siladi et al., 2022), and show sustained suppression at the mRNA level. These same RPGs are suppressed by disruption of the MYC–WDR5 interaction (Thomas et al., 2019), implying that the function of WDR5 at these genes is to recruit MYC, a prominent target of MLL-fusion oncoproteins (Ahmadi et al, 2021). Previously, we posited that suppression of half the RPGs would lead to a ribosomal subunit imbalance (Aho et al., 2019a), but our analysis shows that WINi deplete the entire ribosomal inventory. Part of this depletion is driven by decreased translation of ribosomal mRNAs, although it is also possible that ubiquitin-mediated quality control mechanisms or ribophagy (Zhao et al, 2022) degrade ribosomal proteins under these conditions.

The effects of WINi on ribosomal protein levels are extensive in terms of the number of impacted subunits, but not the magnitude of change, which caps at around a 30% decrease by 72 hours. This cap could be set by the maximal contribution of MYC–WDR5 to the expression of target RPGs, which is two-fold as determined by genetic disruption of the MYC–WDR5 interaction (Thomas et al., 2019), degradation of WDR5 (Siladi et al., 2022), or WIN site blockade. Although modest, it should be noted that a 30% decrease in ribosomal protein levels corresponds to a loss of up to three million ribosomes per cell (Shore & Albert, 2022), and in that light it is not surprising that WINi result in reduced translational efficiencies of about half of all actively-translated mRNAs. Unlike perturbations such as ER stress (Advani & Ivanov, 2019), WINi does not promote preferential translation of stress-responsive mRNAs, although there is some specificity in terms of the translational consequences. Messenger RNAs carrying 5’TOP motifs, for example, are spared from the full translational impact of C6/C16, and we see distinct biological clustering of mRNAs with decreased translation efficiency. Whether these patterns are intrinsic to WINi, or a general response of MV4;11 cells to translational stress, remains to be determined. Overall, we conclude that WINi do not trigger extensive translational reprogramming, but rather act to induce a widespread yet restrained translational choke.

Downstream of ribosomal protein attrition, we observe several responses to WINi that are shared with rRNA inhibitors. Induction of DNA damage, as measured by accumulation of ɣ-H2A.X, is also seen with the RNA polymerase I inhibitor CX-5461 (El Hassouni et al, 2019; Quin et al, 2016), and this has been proposed to be an important and exploitable attribute in terms of its future clinical application (Sanij et al, 2020). The fact that not all rRNA inhibitors induce DNA damage [e.g., (Peltonen et al., 2014)], as well as the action of CX-5461 as a G-quadruplex (G4) stabilizer (Xu et al, 2017), has cast doubt on the contribution of rRNA inhibition versus G4 stabilization to the DNA damage component of CX-5461 activity. But the demonstration that C6/C16 induce ɣ-H2A.X and depend on components of the DDR for full activity, together with a growing appreciation of the susceptibility of rDNA to damage and of the nucleolus as a hub for the DDR (Xuan et al, 2021)—suggests that DNA damage and the response to it may be an inherent part of the mechanism of action of ribosome-directed anti-cancer strategies.

Activation of p53, and p53-dependent cellular inhibition, is another point of convergence of WDR5 and rRNA inhibitors. Not only do we detect activation of p53 target genes in response to WIN site blockade, but we observe synergistic activation of p53 target genes with the BET bromodomain inhibitor mivebresib and suppression of p53 targets upon deletion of RPL22; both of which correlate with cellular sensitivity. We also recover multiple components of the p53 signaling pathway as “resistance” genes in our CRISPR screens, including p53 itself and the splicing factor ZMAT3. Inhibition of rRNA synthesis is thought to activate p53 by generating excess ribosomal proteins that bind to and inactivate MDM2 (Pfister, 2019). It is likely that inhibition of MDM2 contributes to p53 activation in response to C6/C16, as we see a modest increase in p53 levels in MV4;11 cells exposed to WINi, and find that loss of the MDM2 inhibitor p14ARF renders MV4;11 cells less sensitive to C6/C16. Here, however, we also find that there is a second route of p53 activation at work, in which WINi promote accumulation of alternatively spliced isoforms of MDM4 mRNA in which exon 6 is skipped, driving down MDM4 protein levels. Because MDM4 inhibits p53 via proteolysis-independent mechanisms (Francoz et al., 2006), these findings explain how WIN site inhibitors can induce a robust p53 target gene signature in the absence of frank induction of p53 protein. They also point to a dominant role of the MDM4–p53 axis in mediating the response of MLLr cells to WIN site blockade. Whether rRNA inhibition triggers p53 activation by a similar mechanism has, to our knowledge, yet to be reported.

In considering the mechanism through which WINi inactivate MDM4, it is possible that the balance of RPL22L1 and ZMAT3, which are oppositely regulated in response to WIN site inhibition, governs the extent of MDM4 exon 6 inclusion. ZMAT3 is induced in response to p53 activation and promotes skipping of exon 6 in MDM4 (Bieging-Rolett et al., 2020). RPL22L1, in contrast, which is potently suppressed by WIN site inhibitor, promotes exon 6 inclusion (Larionova et al., 2022). Any process that tips the balance in favor of ZMAT3, therefore, would be expected to inhibit MDM4, activate p53, and initiate a feed-forward mechanism that drives ZMAT3 expression and fortifies p53 induction. Induction of ZMAT3 alone should be sufficient to trigger this circuit, but the conspicuous suppression of RPL22L1 by WIN site inhibitor suggests that this may also contribute to the response. Paradoxically, the isoform of RPL22L1 that is suppressed by WIN site blockade, RPL22L1a, is linked in glioblastoma cells to ribosome function, not splicing (Larionova et al., 2022), while the splicing-relevant RPL22L1b isoform is resistant to WINi. One possibility is that, despite its recurrence and prominence, suppression of RPL22L1a by C6/C16 does not contribute to MDM4 suppression, and induction of ZMAT3 is the critical driving factor. Alternatively, RPL22L1a may indeed act to control splicing in MLLr cells, in contrast to what has been reported in glioblastoma. Further investigation is required.

The RPL22–RPL22L1–MDM4 nexus we encountered has been inferred by genome-wide studies of the Cancer Cell Line Encyclopedia [CCLE; (Ghandi et al., 2019)], and our work here demonstrates that loss of RPL22 causes induction of RPL22L1 and promotes inclusion of exon 6 in MDM4. We have no evidence that RPL22 itself plays a role in the response to WINi in otherwise unperturbed MLLr cells. Rather, we suggest that its recovery as a resistance gene in our CRISPR screen is tied to its ability to suppress RPL22L1 expression (O’Leary et al, 2013), and the ectopic effect of its deletion on preventing WINi-induced RPL22L1a decline. That said, RPL22 status is likely to be highly relevant in terms of patient selection criteria. RPL22 is frequently inactivated or deleted in primary cancer samples (Ghandi et al., 2019; Goudarzi & Lindstrom, 2016; Kandoth et al, 2013), as well as 7% of lines in the CCLE (Cao et al, 2017). Unlike other RPGs, mutation or deletion of RPL22 is not associated with mutational inactivation of p53, and indeed there is a strong tendency for wild-type p53 to be retained in RPL22 mutant/deletion lines (Cao et al., 2017; Ghandi et al., 2019). In practical terms, therefore, cancers that retain wild-type p53 but otherwise are mutated/deleted for RPL22, or overexpress RPL22L1, would not be expected to robustly respond to WIN site inhibitor.

Although p53 is important for the action of WINi in MLLr cells, there are likely other stress response mechanisms that mediate cellular inhibition by these agents. A number of p53-independent nucleolar stress responses have been identified [e.g., (Boglev et al, 2013; Jayaraman et al, 2017; Pfister et al, 2015)], including those involving DNA damage response pathways that delay S-phase progression and induce a G2 cell cycle arrest (Sanij et al., 2020). Conversely, we might also expect cells to be able to mount protective responses to WIN site blockade. In this regard, it is curious that four resistance genes identified in our CRISPR screen—UBA6, BIRC6, KCMF1, and UBR4—encode members of a newly-identified BIRC6 ubiquitin-ligase complex (Cervia et al, 2023), the function of which is to prevent aberrant activation of the integrated stress response (ISR). The ISR is a central regulator of protein homeostasis (Costa-Mattioli & Walter, 2020) that drives protective translational reprogramming in response to multiple cellular stresses. There is no indication that the ISR is activated by WINi; indeed the master regulator of ISR, ATF4, is suppressed by C6/C16 at the mRNA and translational levels. But the finding that loss of all four members of the BIRC6 complex blunts the response to WINi implies that ISR activation can be a mechanism through which cells evade the full impact of these agents.

As with most monotherapies, future single agent WINi treatment paradigms are likely to encounter resistance, either by activation of protective responses such as those proposed above, or by mutations in one or more of the resistance genes recovered in our CRISPR screen. Identification of agents that can be used in combination with WINi to increase cancer cell inhibition is thus crucial. Our relatively limited synergy screening identified a number of combinations that should be prioritized for in vivo testing. We found that WINi act synergistically with the BCL-2 inhibitor venetoclax (Souers et al., 2013). This is mechanistically rationalized by our recovery of BAX as a resistance gene, and is noteworthy because venetoclax is an approved therapy for several blood-borne cancers. We also identified notable synergies with experimental agents targeting BET bromodomain family members, DOT1L, and PRMT5. The combination with mivebresib is rationalized based on identification of BRD3 as a sensitizing gene, and likely results from the ability of C16 and mivebresib to inhibit distinct sets of genes connected to translation, the impact of which is to enhance p53 induction. BET bromodomain inhibitors have struggled in clinical trials due to dose-limiting toxicities (Shorstova et al, 2021), but their combination with WINi could form the basis of a more effective therapy with less side effects. Moreover, given the mechanism the underlying synergy between C16 and BET inhibitors, we would expect this combination to be effective in other wild-type p53 cancer settings where WINi are active, such as neuroblastoma (Bryan et al., 2020) and rhabdoid tumors (Florian et al., 2022). Expanded synergy screening is needed to identify and understand the full spectrum of combination approaches that could be used to ultimately enhance and extend the clinical utility of WIN site inhibitors.

Supporting information

eLife Assessment

The manuscript analyses the mechanism of action of small compounds that target the WDR5 chromatin protein for tumor therapy. They study a class of compounds termed WIN site inhibitors and show that they affect the expression of ribosome protein genes (RPGs) and cause ribosomal stress, ultimately activating the TP53 tumor suppressor protein. The manuscript provides important insight into the mode of action of these interesting compounds. The evidence is convincing, with a few exceptions detailed in the reviewers' comments.

Reviewer #1 (Public Review):

Building on previous work from the Tansey lab, here Howard et al. characterize transcriptional and translational changes upon WIN site inhibition of WDR5 in MLL-rearranged cancer cells. They first analyze whether C16, a newer generation compound, has the same cellular effects as C6, an early generation compound. Both compounds reduce the expression of WDR5-bound RPGs in addition to the unbound RPG RPL22L1. They then investigate differential translation by ribo-seq and observe that WIN site inhibition reduces the translational RPGs and other proteins related to biomass accumulation (spliceosome, proteasome, mitochondrial ribosome). Interestingly, this reduction adds to the transcriptional changes and is not limited to RPGs whose promoters are bound by WDR5. Quantitative proteomics at two-time points confirmed the downregulation of RPGs. Interestingly, the overall effects are modest, but RPL22LA is strongly affected. Unexpectedly, most differentially abundant proteins seem to be upregulated 24 h after C6 (see below). A genetic screen showed that loss of p53 rescues the effect of C6 and C16 and helped the authors to identify pathways that can be targeted by compounds together with WIN site inhibitors in a synergistic way. Finally, the authors elucidated the underlying mechanisms and analyzed the functional relevance of the RPL22, RPL22L1, p53, and MDM4 axis.

While this work is not conceptually new, it is an important extension of the observations of Aho et al. The results are clearly described and, in my view, very meaningful overall.

Major points:

1. The authors make statements about the globality/selectivity of the responses in RNA-seq, ribo-seq, and quantitative proteomics. However, as far as I can see, none of these analyses have spike-in controls. I recommend either repeating the experiments with a spike-in control or carefully measuring transcription and translation rates upon WIN site inhibition and normalizing the omics experiments with this factor.

2. Why are the majority of proteins upregulated in the proteomics experiment after 24 h in C6 (if really true after normalization with general protein amount per cell)? This is surprising and needs further explanation.

3. The description of the two CRISPR screens (GECKO and targeted) is a bit confusing. Do I understand correctly that in the GECKO screen, the treated cells are not compared with non-treated cells of the same time point, but with a time point 0? If so, this screen is not very meaningful and perhaps should be omitted. Also, it is unclear to me what the advantages of the targeted screen are since the targets were not covered with more sgRNAs (data contradictory: 4 or 10 sgRNAs per target?) than in Gecko. Also, genome-wide screens are feasible in culture for multiple conditions. Overall, I find the presentation of the screening results not favorable.

4. Can Re-expression of RPL22 rescue the growth arrest of C6?.

Reviewer #2 (Public Review):

Summary:

The manuscript by Howard et al reports the development of high-affinity WDR5-interaction site inhibitors (WINi) that engage the protein to block the arginine-dependent engagement with its partners. Treatment of MLL-rearranged leukemia cells with high-affinity WINi (C16) decreases the expression of genes encoding most ribosomal proteins and other proteins required for translation. Notably, although these targets are enriched for WDR5-ChIP-seq peaks, such peaks are not universally present in the target genes. High concordance was found between the alterations in gene expression due to C16 treatment and the changes resulting from treatment with an earlier, lower affinity WINi (C6). Besides protein synthesis, genes involved in DNA replication or MYC responses are downregulated, while p53 targets and apoptosis genes are upregulated. Ribosome profiling reveals a global decrease in translational efficiency due to WINi with overall ribosome occupancies of mRNAs ~50% of control samples. The magnitude of the decrements of translation for most individual mRNAs exceeds the respective changes in mRNA levels genome-wide. From these results and other considerations, the authors hypothesize that WINi results in ribosome depletion. Quantitative mass spec documents the decrement in ribosomal proteins following WINi treatment along with increases in p53 targets and proteins involved in apoptosis occurring over 3 days. Notably, RPL22L1 is essentially completely lost upon WINi treatment. The investigators next conduct a CRISPR screen to find moderators and cooperators with WINi. They identify components of p53 and DNA repair pathways as mediators of WINi-inflicted cell death (so gRNAs against these genes permit cell survival). Next, WINi are tested in combination with a variety of other agents to explore synergistic killing to improve their expected therapeutic efficacy. The authors document the loss of the p53 antagonist MDM4 (in combination with splicing alterations of RPL22L1), an observation that supports the notion that WINi killing is p53-mediated.

Strengths:

This is a scientifically very strong and well-written manuscript that applies a variety of state-of-the art molecular approaches to interrogate the role of the WDR5 interaction site and WINi. They reveal that the effects of WINi seem to be focused on the overall synthesis of protein components of the translation apparatus, especially ribosomal proteins-even those that do not bind WDR5 by ChIP (a question left unanswered is how much the WDR5-less genes are nevertheless WINi targeted). They convincingly show that disruption of the synthesis of these proteins is accompanied by DNA damage inferred by H2AX-activation, activation of the p53-pathway, and apoptosis. Pathways of possible WINi resistance and synergies with other anti-neoplastic approaches are explored. These experiments are all well-executed and strongly invite more extensive pre-clinical and translational studies of WINi in animal studies. The studies also may anticipate the use of WINi as probes of nucleolar function and ribosome synthesis though this was not really explored in the current manuscript.

Weaknesses:

A mild deficiency in the current manuscript is the absence of cell biological methods to complement the molecular biological and biochemical approaches so ably employed. Some microscopic observations and confirmation of nucleolar dysfunction and DNA damage would be reassuring.

Author response:

The following is the authors’ response to the original reviews.

We are grateful to the Editors for overseeing the review of our manuscript, and to the two reviewers for their thoughtful comments and suggestions for how it can be improved.

I submit at this time a revision, as well as a detailed response (below) to each of the points raised in the first round of review.

We feel the manuscript has been significantly improved by taking the reviewers' comments to heart. In a nutshell, we added new key pieces of data (impact of WIN site inhibition on global translation, rRNA production, as well as the requested cell biology analyses showing nucleolar stress), new analyses of the proteomics to counter potential concerns with normalization, and expanded/revised verbiage in key areas to clarify parts of the text that were confusing or problematic. The main figures have not changed; all new material is included in supplements to figures 2 and 3.

Public Reviews

Reviewer #1 (Public Review):

Building on previous work from the Tansey lab, here Howard et al. characterize transcriptional and translational changes upon WIN site inhibition of WDR5 in MLL-rearranged cancer cells. They first analyze whether C16, a newer generation compound, has the same cellular effects as C6, an early generation compound. Both compounds reduce the expression of WDR5-bound RPGs in addition to the unbound RPG RPL22L1. They then investigate differential translation by ribo-seq and observe that WIN site inhibition reduces the translational RPGs and other proteins related to biomass accumulation (spliceosome, proteasome, mitochondrial ribosome). Interestingly, this reduction adds to the transcriptional changes and is not limited to RPGs whose promoters are bound by WDR5. Quantitative proteomics at two-time points confirmed the downregulation of RPGs. Interestingly, the overall effects are modest, but RPL22LA is strongly affected. Unexpectedly, most differentially abundant proteins seem to be upregulated 24 h after C6 (see below). A genetic screen showed that loss of p53 rescues the effect of C6 and C16 and helped the authors to identify pathways that can be targeted by compounds together with WIN site inhibitors in a synergistic way. Finally, the authors elucidated the underlying mechanisms and analyzed the functional relevance of the RPL22, RPL22L1, p53, and MDM4 axis.

While this work is not conceptually new, it is an important extension of the observations of Aho et al. The results are clearly described and, in my view, very meaningful overall.

Major points:

(1) The authors make statements about the globality/selectivity of the responses in RNA-seq, ribo-seq, and quantitative proteomics. However, as far as I can see, none of these analyses have spike-in controls. I recommend either repeating the experiments with a spike-in control or carefully measuring transcription and translation rates upon WIN site inhibition and normalizing the omics experiments with this factor.

The reviewer is correct that we did not include spike-in controls in our omics experiments. We would like to emphasize that none of the omics data in this manuscript have been processed in unorthodox ways, and that the major conclusions each have independent corroborating data.

The selectivity in RPG suppression observed in RNA-Seq, for example, is supported by results from our target engagement (QuantiGene) assays; suppression of RPL22L1 mRNA levels is supported by quantitative and semi-quantitative RT-PCR, by western blotting, and by the results of our proteomic profiling; alternative splicing (and expression) of MDM4—and its dependency on RPL22—is also backed up by similar RT-PCR and western blotting data. The same applies for alternative splicing of RPL22L1.

That said, we do appreciate the point the reviewer is making here, and have done our best to respond. We do not think it is a prudent investment in resources to repeat the numerous omics assays in the manuscript. We also considered normalizing for bulk transcription and translation rates as suggested, but it is not clear in practice how this would be done, and it could introduce additional variables and uncertainties that may skew the interpretation of results. Instead, to respond to this comment, we made the following changes to the manuscript:

(1) We now explicitly state, for all omics assays, that spike-in controls were not included. These statements will prompt the reader to make their own assessment of the robustness of each of our findings and interpretations.

(2) We have added new data to the manuscript (Figure 2—figure supplement 1A–B) measuring the impact of C6 and C16 on bulk translation using the OPP labeling method. These new data demonstrate that WIN site inhibitors induce a progressive yet modest decline in protein synthesis capacity. At 24 hours, there is no significant effect of either agent on protein synthesis levels. By 48 hours, a small but significant effect is observed, and by 96 hours translation levels are ~60% of what they are in vehicle-treated control cells. These new data are important because they support the idea that normalization has not blunted the responses we observe—the magnitude of the effects are consistent between the different assays and tend to cap out at two-fold in terms of RPG suppression, translation efficiency, ribosomal protein levels, and protein synthesis capacity.

(3) We have included additional analysis regarding the LFQMS, as described below, that specifically addresses the issue of normalization in our proteomics experiments.

(2) Why are the majority of proteins upregulated in the proteomics experiment after 24 h in C6 (if really true after normalization with general protein amount per cell)? This is surprising and needs further explanation.

The reviewer is correct in noting that (by LFQMS) ~700 proteins are induced after 24 hours of treatment of MV4:11 cells with C16 (not C6, as stated). The reviewer would like us to examine whether this apparent increase in proteins is a normalization artifact. In response to this comment, we have made the following changes to the manuscript:

(1) Our new OPP labeling experiments (Figure 2—figure supplement 1A–B) show that there is no significant reduction in overall protein synthesis following 24 hours of C16 treatment. In light of this finding, it is unlikely that normalization artifacts, resulting from diminution of the pool of highly abundant proteins, create the appearance of these 700 proteins being induced. We now explicitly make this point in the text.

(2) We now clarify in the methods how we seeded identical numbers of cells for DMSO and C16-treated cultures in these experiments, and—consistent with our finding that WIN site inhibitors have little if any effect on protein synthesis or proliferation at the 24 hour timepoint— extracted comparable amounts of proteins from these two treatment conditions (DMSO: 344.75 ± 21.7 µg; C16: 366.50 ± 15.8 µg; [Mean ± SEM]).

(3) We now include in Figure 3—figure supplement 1A a plot showing the distribution of peptide intensities for each protein detected in each run of LFQMS before and after equal median normalization. This new analysis reveals that the distribution of intensities is not appreciably changed via normalization. Specifically, there is not a reduction in peptide intensities in the unnormalized data from 24 hours of C16 treatment that is reversed or tempered by normalization. This analysis provides further support for the notion that the increase we observe is not a normalization artifact.

(4) We now include in Figure 3—figure supplement 1B–D a set of new analyses examining the relationship between the initial intensity of proteins in DMSO control samples (a crude proxy for abundance) versus the fold change in response to WIN site inhibitor. This analysis shows that we have as many "highly abundant" (10th decile) proteins increasing as we do decreasing in response to WINi. Thus, it appears as though the wholesale clearance of highly abundant proteins from the cell is not occurring at this early treatment timepoint. In addition, this analysis also shows that ribosomal proteins (RP) are generally the most abundant, most suppressed, proteins and that their fold-change at the protein level at 24 hours is less than two-fold, consistent again with the magnitude of transcriptional effects of C16, as measured by RNA-Seq and QuantiGene. The fact that the drop in RP levels is consistent with expectations based on other analyses provides further empirical support for the notion that protein levels inferred from LFQMS are authentic and not skewed by global changes in the proteome.

The increase in proteins at this time point, we argue, is thus most likely genuine. It is not surprising that—at a timepoint at which protein synthesis is unaffected—several hundred proteins are induced by a factor of two. How this occurs, we do not know. It may be a transient compensatory mechanism, or it may be an early part of the active response to WIN site inhibitors. Lest the reader be confused by this finding, we have now added text to this section of the manuscript discussing and explaining the phenomenon in more detail.

(3) The description of the two CRISPR screens (GECKO and targeted) is a bit confusing. Do I understand correctly that in the GECKO screen, the treated cells are not compared with nontreated cells of the same time point, but with a time point 0? If so, this screen is not very meaningful and perhaps should be omitted. Also, it is unclear to me what the advantages of the targeted screen are since the targets were not covered with more sgRNAs (data contradictory: 4 or 10 sgRNAs per target?) than in Gecko. Also, genome-wide screens are feasible in culture for multiple conditions. Overall, I find the presentation of the screening results not favorable.

In essence, this is a single screen performed in two tiers. In Tier 1, we screened a complete GECKO library (six sgRNA/gene) with the earliest generation (less potent) inhibitor C6, and compared sgRNA representation against the time zero population. This screen would reveal sgRNAs that are specifically associated with response to C6, as well as those that are associated with general cell fitness and viability. We then identified genes connected to these sgRNAs, removed those that are pan essential, and built a custom library for the second tier using sgRNAs from the Brunello library (four sgRNA/gene). We then screened this custom library with both C6 and the more potent inhibitor C16, this time against DMSO-treated cells from the same timepoint.

We acknowledge that this is not the most streamlined setup for a screen. But our intention was to compare two inhibitors (C6 and C16) and identify high confidence 'hits' that are disconnected from general cell viability, rather than generate an exhaustive list of all genes that, when disrupted, skew the response to WIN site inhibitor. The final result of this screen (Figure 4E) is a gene list that has been validated with two chemically distinct WIN site inhibitors and up to 10 unique sgRNAs per gene. We may not have captured every gene that can modulate response to WIN site inhibitor, but those appearing in Figure 4E are highly validated.

To answer the reviewer's specific questions: (i) we cannot omit the Tier 1 screen because then there would be no rationale for what was screened in the second Tier; and (ii) the advantage of the custom Tier 2 library is that it allowed us to screen hits from the Tier 1 screen with four completely independent sgRNAs. Although there are not more sgRNAs for each gene in the Tier 2 versus the Tier 1 library, these sgRNAs are different and thus, for C6 at least, hits surviving both screens were validated with up to 10 unique sgRNAs.

We apologize that the description of the CRISPR screens was not clearer, and have reworked this section of the manuscript to make our intent and our actions clearer.

(4) Can Re-expression of RPL22 rescue the growth arrest of C6?.

We have not attempted to complement the RPL22 knock out. But we do note that evidence supporting the idea that loss of RPL22 confers resistance to WIN site inhibitor is strong—six (out of six) sgRNAs against RPL22 were significantly enriched in the Tier 1 screen, and independent knock out of RPL22 with the Synthego multi-guide system in MV4;11 and MOLM13 cells increases the GI50 for C16.

Reviewer #2 (Public Review):

Summary:

The manuscript by Howard et al reports the development of high-affinity WDR5-interaction site inhibitors (WINi) that engage the protein to block the arginine-dependent engagement with its partners. Treatment of MLL-rearranged leukemia cells with high-affinity WINi (C16) decreases the expression of genes encoding most ribosomal proteins and other proteins required for translation. Notably, although these targets are enriched for WDR5-ChIP-seq peaks, such peaks are not universally present in the target genes. High concordance was found between the alterations in gene expression due to C16 treatment and the changes resulting from treatment with an earlier, lower affinity WINi (C6). Besides protein synthesis, genes involved in DNA replication or MYC responses are downregulated, while p53 targets and apoptosis genes are upregulated. Ribosome profiling reveals a global decrease in translational efficiency due to WINi with overall ribosome occupancies of mRNAs ~50% of control samples. The magnitude of the decrements of translation for most individual mRNAs exceeds the respective changes in mRNA levels genome-wide. From these results and other considerations, the authors hypothesize that WINi results in ribosome depletion. Quantitative mass spec documents the decrement in ribosomal proteins following WINi treatment along with increases in p53 targets and proteins involved in apoptosis occurring over 3 days. Notably, RPL22L1 is essentially completely lost upon WINi treatment. The investigators next conduct a CRISPR screen to find moderators and cooperators with WINi. They identify components of p53 and DNA repair pathways as mediators of WINi-inflicted cell death (so gRNAs against these genes permit cell survival). Next, WINi are tested in combination with a variety of other agents to explore synergistic killing to improve their expected therapeutic efficacy. The authors document the loss of the p53 antagonist MDM4 (in combination with splicing alterations of RPL22L1), an observation that supports the notion that WINi killing is p53-mediated.

Strengths:

This is a scientifically very strong and well-written manuscript that applies a variety of state-ofthe art molecular approaches to interrogate the role of the WDR5 interaction site and WINi. They reveal that the effects of WINi seem to be focused on the overall synthesis of protein components of the translation apparatus, especially ribosomal proteins-even those that do not bind WDR5 by ChIP (a question left unanswered is how much the WDR5-less genes are nevertheless WINi targeted). They convincingly show that disruption of the synthesis of these proteins is accompanied by DNA damage inferred by H2AX-activation, activation of the p53pathway, and apoptosis. Pathways of possible WINi resistance and synergies with other antineoplastic approaches are explored. These experiments are all well-executed and strongly invite more extensive pre-clinical and translational studies of WINi in animal studies. The studies also may anticipate the use of WINi as probes of nucleolar function and ribosome synthesis though this was not really explored in the current manuscript.

Weaknesses:

A mild deficiency in the current manuscript is the absence of cell biological methods to complement the molecular biological and biochemical approaches so ably employed. Some microscopic observations and confirmation of nucleolar dysfunction and DNA damage would be reassuring.

We thank the reviewer for their comments. We agree that an absence of cell biological methods was a deficiency in the original manuscript. In response to this comment, we have now added immunofluorescence (IF) analyses, examining the impact of C16 on nucleolar integrity and nucleophosmin (NPM1) distribution (Figure 3—figure supplement 4). These new data clearly show that C16 induces nucleolar stress at 72 hours—as measured by the redistribution of NPM1 from the nucleolus to the nucleoplasm. These new data fill an important gap in the story, and we are grateful to the reviewer for prompting us to perform these experiments.

As part of the above study, we also probed for gamma-H2AX, expecting that we may see some signs of accumulation in the nucleoli (see comment #4 from Reviewer #2, below). We did not observe this response. Importantly, however, we did see that gamma-H2AX staining occurs only in what are overtly apoptotic cells. This is an important finding, because we had previously speculated that the induction of gamma-H2AX observed by Western blotting reflected part of a bona-fide response to DNA damage elicited by WIN site inhibitors. Instead, the IF data now leads us to conclude that this signal simply reflects the established fact that WIN site inhibitors induce apoptosis in this cell line (Aho et al., 2019). In response to this new finding, we have added additional discussion to the text and have removed or de-emphasized the potential contribution of DNA damage to the mechanism of action of WDR5 WIN site inhibitors. Again, we are grateful for this comment as it has prevented us from continuing to report/pursue erroneous observations.

Recommendations for the authors

Reviewer #1 (Recommendations For The Authors):

There is a typo in "but are are linked to mRNA instability when translation is inhibited".

Thank you for catching this typo. It has now been corrected.

Reviewer #2 (Recommendations For The Authors):

(1) The authors report that WINi initially (at 24 hrs) increases the expression of most proteins while decreasing ribosomal proteins, but at 72 hours all proteins are depressed. The transient bump-up of non-translation-related proteins seems odd. A simple resolution to this somewhat strange observation is that there is no real increase in the other proteins, but because of the loss of a large fraction of the most abundant cellular proteins (the ribosomal proteins), the relative fraction of all other proteins is increased; that is, the increase of non-ribosomal proteins may be an artifact of normalization to a lower total protein content. Can this be explored?

We are grateful to the reviewer for this comment. We have tried our best to respond, as detailed above in response to Reviewer #1 Public Comment #2.

(2) It would be really nice to assess nucleolar status microscopically. Do nucleoli get bigger? Smaller? Do they have abnormal morphology? Is there nucleolar stress? What happens to rRNA synthesis and processing?

We agree and thank the reviewer for raising this point. As noted in our response to Reviewer #2, above, we have included new IF that shows: (i) no obvious effect on nucleolar integrity, (ii) redistribution of NPM1 to the nucleoplasm (indicative of nucleolar stress), and (iii) induction of gamma-H2AX staining in apoptotic cells (indicative of apoptosis).

Additionally, in response to this comment, we also looked at the impact of WIN site inhibitors on rRNA synthesis, using AzCyd labeling. These new data appear in Figure 3—figure supplement 3. Interestingly, these new data show that there is a progressive decline in rRNA synthesis, and that by 96 hours of treatment levels of both 18S and 28S rRNAs are reduced— again by about a factor of two. Our interpretation of this finding is that in response to the progressive decline in RPG transcription there is a secondary decrease in rRNA synthesis. This result is perhaps not surprising, but it does again add an important missing piece to our characterization of WIN site inhibitors and is further support for the concept that inhibition of ribosome production is a dominant part of the response to these agents.

(3) The WINi elicited DNA damage is incompletely characterized, rather it is inferred from H2AX activation. Comet assays would help to confirm such damage.

As noted in our response to Reviewer #2, our original inference of DNA damage, prompted by gamma-H2AX activation, is erroneous, and due instead to the ability of WIN site inhibitors to induce apoptosis. We thus did not pursue comet assays, etc., and removed discussion of potential DNA damage from the manuscript.

(4) Staining and microscopic observation of H2AX would be very useful. Is the WINi provoked DNA damage nucleolar-localized? Does the deficiency of ribosomal proteins lead to localized genotoxic nucleolar stress - or alternatively does the paucity of ribosomes and decreased translation lead to imbalances in other cellular pathways, perhaps including some involved in overall genome maintenance which would provoke more global DNA damage and H2AX staining, not limited to the nucleolus.

Again, please see our response to the Public Comment from Reviewer #2.

(5) It would be important to assess the influence and effects of WINi on some p53 mutant, p53-/- and p53 wild-type cell lines. Given their prevalence, p53 status may be expected to alter WINi efficacy.

The issue of how p53 status impacts the response to WINi is interesting and important, but we feel this is beyond the scope of the current manuscript. It is likely that many factors contribute to the response of cancer cells to these agents, and thus simply surveying some cancer lines for their response and linking this to their p53 status is unlikely to be very informative. Making definitive statements about the contribution of p53, and the differences between wild-type, lossof-function mutants, gain of function mutants, and null mutants will require more extensive analyses and is fertile territory for future studies, in our opinion.

Version 2

Abstract

Summary

The chromatin-associated protein WD Repeat Domain 5 (WDR5) is a promising target for cancer drug discovery, with most efforts blocking an arginine-binding cavity on the protein called the "WIN" site that tethers WDR5 to chromatin. WIN site inhibitors (WINi) are active against multiple cancer cell types in vitro, the most notable of which are those derived from MLL-rearranged (MLLr) leukemias. Peptidomimetic WINi were originally proposed to inhibit MLLr cells via dysregulation of genes connected to hematopoietic stem cell expansion. Our discovery and interrogation of small molecule WIN site inhibitors, however, revealed that they act in MLLr cell lines to suppress ribosome protein gene (RPG) transcription, induce nucleolar stress, and activate p53. Because there is no precedent for an anti-cancer strategy that specifically targets RPG expression, we took an integrated multi-omics approach to further interrogate the mechanism of action of WINi in MLLr cancer cells. We show that WINi induce depletion of the stock of ribosomes, accompanied by a broad yet modest translational choke and changes in alternative mRNA splicing that inactivate the p53 antagonist MDM4. We also show that WINi are synergistic with agents including venetoclax and BET-bromodomain inhibitors. Together, these studies reinforce the concept that WINi are a novel type of ribosome-directed anti-cancer therapy and provide a resource to support their clinical implementation in MLLr leukemias and other malignancies.

Introduction

WDR5 is a highly-conserved protein that moonlights in a variety of functions in the nucleus. It rose to prominence as a component of epigenetic writer complexes that deposit histone H3 lysine 4 (H3K4) methylation (Guarnaccia and Tansey, 2018), but was subsequently found to act outside these complexes to facilitate the integrity of the mitotic spindle (Ali et al., 2017), bookmark genes for reactivation after mitosis (Oh et al., 2020), and promote transcription of a subset of ribosomal protein genes [(RPGs); (Bryan et al., 2020)] via recruitment of the oncoprotein transcription factor MYC to chromatin (Thomas et al., 2019). WDR5 is also frequently overexpressed in cancer, where its overexpression correlates with aggressive disease and poor clinical outcomes (Guarnaccia and Tansey, 2018). Accordingly, WDR5 is an auspicious target for inhibition in a range of malignancies including MLL-rearranged (MLLr) leukemias (Cao et al., 2014, Aho et al., 2019a), MYC-driven cancers (Aho et al., 2019b), C/EBPα-mutant leukemias (Grebien et al., 2015), p53 gain-of-function cancers (Zhu et al., 2015), neuroblastomas (Bryan et al., 2020), rhabdoid tumors (Florian et al., 2022), and metastatic breast cancers (Cai et al., 2022).

Although WDR5 PROTACs have been described (Yu et al., 2021, Li et al., 2022, Yu et al., 2023), safety concerns over destroying a pan-essential protein such as WDR5 (Siladi et al., 2022) means that most drug discovery efforts have focused on small molecule inhibition of key binding sites on the protein. Some initiatives target a hydrophobic cleft on WDR5 known as the "WDR5-binding motif" (WBM) site (Macdonald et al., 2019, Chacon Simon et al., 2020) that contacts MYC (Thomas et al., 2015). But the majority of efforts target the "WDR5-interaction" (WIN) site of WDR5 (Guarnaccia and Tansey, 2018)— an arginine binding cavity that tethers WDR5 to chromatin (Aho et al., 2019a) and makes contact with partner proteins carrying an arginine-containing "WIN" motif (Guarnaccia et al., 2021). Multiple WIN site inhibitors (WINi) have been described (Bolshan et al., 2013, Karatas et al., 2013, Senisterra et al., 2013, Cao et al., 2014, Grebien et al., 2015, Li et al., 2016, Karatas et al., 2017, Wang et al., 2018, Aho et al., 2019a, Tian et al., 2020, Chen et al., 2021a, Chen et al., 2021b), including those that are orally bioavailable and have anti-tumor activity in vivo (Chen et al., 2021b, Teuscher et al., 2023). Given the intense interest in developing WINi for cancer therapy, and the rapid pace of improvement in these molecules, it is likely that WIN site inhibitors will be ready for clinical vetting in the near future.

That said, controversy remains regarding the mechanism of action of WIN site inhibitors, even in the context of MLLr leukemias, where there is strong empirical support for their utility (Weissmiller et al., 2024). MLL-rearranged leukemias are defined by translocation of one copy of MLL1—a gene that encodes one of six MLL/SET proteins that are the catalytic subunits of the histone methyltransferase (HMT) complexes responsible for H3K4 methylation (Guarnaccia and Tansey, 2018). The near universal retention of a pristine copy of MLL1 in these cancers led to the idea that MLLr leukemias depend on wild-type MLL1 to support the activity of oncogenic MLL1-fusion oncoproteins (Thiel et al., 2010)—a function in turn that depends on insertion of a low affinity WIN motif within MLL1 into the WIN site of WDR5 (Alicea-Velazquez et al., 2016). Consistent with this notion, early peptidomimetic WINi are active against MLLr leukemia cells in vitro and are reported to suppress levels of H3K4 methylation at canonical MLL1-fusion target genes such as the HOXA loci, causing cellular inhibition through a combination of differentiation and apoptosis (Cao et al., 2014). Subsequently, however, wild-type MLL1 was shown to be dispensable for transformation by MLL-fusion oncoproteins (Chen et al., 2017), and our analysis of picomolar small molecule WINi revealed that they act in MLLr cells without inducing significant changes in the expression of HOXA genes or levels of H3K4 methylation (Aho et al., 2019a). Instead, WIN site inhibitors displace WDR5 from chromatin and directly suppress the transcription of ∼50 genes, the majority of which are connected to protein synthesis, including half the cohort of RPGs. We also found that WINi provoke nucleolar stress and induce p53-dependent cell death. Based on our findings, we proposed that WIN site inhibitors kill MLLr cells via depletion of part of the ribosome inventory that induces apoptosis via a ribosome biogenesis stress response.

The concept of ribosome-directed cancer therapies is not new (Laham-Karam et al., 2020, Temaj et al., 2022). Besides mTOR and translational inhibitors, one of the most prevalent strategies in this realm is inhibition of ribosomal RNA (rRNA) production or processing, which is a feature of both existing chemotherapies such as platinum-containing compounds (Bruno et al., 2017), as well as newer targeted RNA polymerase I inhibitors (Drygin et al., 2011, Peltonen et al., 2014). Although these agents exert their anti-cancer effects through multiple mechanisms (Laham-Karam et al., 2020), they are generally thought to disrupt the stoichiometry of RNA and protein components of the ribosome, leading to an excess of ribosomal proteins that inactivate MDM2 to induce p53-dependent cancer cell death. The paradigm we developed for WINi is modeled after that of rRNA inhibitors, although it is important to note that a significant point of divergence from rRNA poisons is that in this model WINi induce p53 not by promoting ribosomal protein accumulation, but by causing a selective imbalance in the ribosome subunit inventory. How such an imbalance could lead to p53 induction, as well as other consequences it may have on cellular processes, remains unknown.

Fortifying understanding of the mechanism of action of WINi in MLLr cancer cells is key to their clinical implementation. At present, there is no precedent for the mechanism we propose, no understanding of the impact of selective ribosome subunit depletion on translation or other tumor-relevant processes, and no expectations for how resistance to WINi could emerge or how their anti-tumor actions could be made more effective. To ameliorate these deficiencies, we took an integrated multiomics approach, combining transcriptional and translational profiling with genome-wide CRISPR screening to probe WINi action in MLLr cells. Our studies show that although the transcriptional effects of WINi on ribosome subunit expression are confined to those RPGs directly regulated by WDR5, effects at the protein level are not, and WIN site inhibition leads to diminution of the entire stock of cytosolic ribosomes. Ribosome subunit attrition is accompanied by a broad translational choke, induction of nucleolar stress, and activation of p53—driven in large part via RPL22-dependent alternative splicing of the p53 antagonist MDM4. We also show that WINi are synergistic with approved and targeted agents including venetoclax and BET-bromodomain inhibitors. Collectively, these findings solidify a novel mechanism of action for WIN site inhibitors in MLLr cells and highlight a path for their optimal clinical implementation.

Results

Impact of WINi on the transcriptome of MLLr cancer cells

Our model for the action of WINi in MLLr leukemia cells is based on analysis of two early generation compounds (Aho et al., 2019a): C3 (Kd = 1.3 nM) and C6 (Kd = 100 pM). Subsequently (Tian et al., 2020), we discovered more potent molecules such as C16 (Kd < 20 pM) that have not been extensively profiled. To determine if improvements in the potency of WINi have resulted in divergent activities, we first compared C6 with C16 (Figure 1A). Both molecules bind the WIN site of WDR5 (Figure 1B and Figure 1—figure supplement 1A), but differ in affinity due to a bicyclic dihydroisoquinolinone core that locks C16 into a favorable binding conformation (Figure 1C). Consistent with its higher affinity, C16 is ∼20 times more potent than C6 in inhibiting MV4;11 (MLL–AF4) and MOLM13 (MLL–AF9) leukemia lines (Figure 1—figure supplement 1B–C): In MV4;11 cells, for example, the GI50 for C6 is 1 μM, compared to 46 nM for C16. These differences in potency are reflected at the level of RPG suppression. Using a target engagement assay (Florian et al., 2022) that measures transcript levels from seven RPGs—five (RPS14, RPS24, RPL26, RPL32, and RPL35) that are always bound by WDR5 and two (RPS11 and RPS14) that are never bound—we observe that maximal suppression of RPG transcripts occurs at ∼2 μM for C6 and ∼100 nM for C16 in MV4;11 (Figure 1D) and MOLM13 (Figure 1—figure supplement 1D) cells. To functionally compare these two inhibitors of different potencies, we used these RPG-normalized doses in all our subsequent studies.

Impact of WINi on the transcriptome of MLLr cancer cells.

(A) Chemical structures of C6 and C16. (B) Crystal structures of C6 or C16 bound to the WIN Site of WDR5 with electrostatic surfaces mapped [PDB IDs: 6E23 (Aho et al., 2019a); 6UCS (Tian et al., 2020)]. The image shows a close-up view of the WIN site. (C) Superimposed WIN site-binding conformations of C6 (green) and C16 (blue). (D) Transcript levels as determined by QuantiGene™ analysis of representative WDR5-bound (color) or non-bound (grayscale) ribosomal protein genes in MV4;11 cells treated with a serial dilution range of either C6 (left) or C16 (right) and relative to DMSO-treated cells (n = 2-3; Mean ±SEM). Vertical dashed line indicates either 2 µM C6 (left) or 100 nM C16 (right). (E) Number of genes with significantly (FDR < 0.05) altered transcript levels following treatment of MV4;11 cells with C6 (2 µM) or C16 (100 nM) for 48 hours, as determined by RNA-Seq (n = 3). See Figure 1—source data 1 for complete output of RNA-Seq analysis. (F) Comparison of gene expression changes elicited by C6 (x-axis) and C16 (y-axis), represented as Log2 fold change (FC) compared to DMSO. WDR5-bound genes are colored red. Locations of RPL22L1 and ZMAT3 are indicated. (G) Overlap of genes with decreased (left) or increased (right) transcript levels in MV4;11 cells treated with C6 or C16. (H) Gene Set Enrichment Analysis (GSEA) showing the distribution of genes suppressed in MV4;11 cells in response to C6 (left) or C16 (right) against the list of all genes bound by WDR5 in those cells (Aho et al., 2019a). NES = normalized enrichment score. (I) Enrichment analysis of genes suppressed (left) or induced (right) by C6 or C16 in MV4;11 cells. KEGG and Hallmark.MSigDB pathways are shown. Fold enrichment of indicated pathways is presented on the x-axis, the number of genes is shown in italics in each bar, and colors represent - Log10 FDR. See Figure 1—source data 2 for additional GSEA (Hallmark) and ORA (Hallmark) analyses of differentially expressed genes. (J) Transcript level changes in WDR5-bound (left) and non-bound (right) RPGs elicited by C6 (top) or C16 (bottom).

We performed RNA-sequencing (RNA-Seq) on MV4;11 cells treated for 48 hours with DMSO, C6, or C16 (Figure 1—figure supplement 1E and Figure 1—source data 1). Spike-in controls were not included. Both compounds elicit thousands of gene expression changes (Figure 1E), a majority of which are less than two-fold in magnitude (Figure 1—figure supplement 1F). We had previously performed RNA-Seq on MV4;11 cells treated with 2 μM C6 and observed just ∼75 induced and ∼460 reduced transcripts (Aho et al., 2019a). In this earlier work, however, increased variance among replicates made it difficult for as many small gene expression changes to reach statistical significance as in the current study (Figure 1—figure supplement 1G). Comparing the new RNA-Seq datasets, we observe similar effects of C6 and C16 on the MV4;11 cell transcriptome (Figure 1F), with more than 80% of the transcripts altered by C16 altered in the same direction by C6 (Figure 1G). In both cases, suppressed genes are enriched in those bound by WDR5 in MV4;11 cells (Figure 1H); although a majority of transcriptional changes occur at loci bereft of detectable WDR5 binding (Figure 1F). For both compounds, expression of genes connected to protein synthesis, the cell cycle, DNA replication, mTORC signaling, and MYC are reduced, while expression of those connected to chemokine signaling, apoptosis and p53 are induced (Figure 1I and Figure 1—source data 2). Indeed, ∼90 "consensus" p53 target genes (Fischer, 2017) are induced by C6/C16 (Figure 1—figure supplement 2A–B), including the tumor suppressor ZMAT3 [Figure 1F; (Bieging-Rolett et al., 2020)]. Finally, we note that for both compounds, the transcriptional effects on RPG expression are almost entirely confined to those RPGs bound by WDR5 (Figure 1J). The conspicuous exception to this trend is RPL22L1—a paralog of RPL22—mRNA levels which are strongly reduced by C6/C16 (Figure 1F and Figure 1J). This feature of the response is not confined to MLLr cells, as RPL22L1 expression is also decreased by WINi in sensitive rhabdoid tumor cell lines (Florian et al., 2022), but not in the insensitive K562 leukemia line (Bryan et al., 2020) (Figure 1—figure supplement 2C). Despite not being a direct WDR5 target gene, therefore, RPL22L1 expression is recurringly suppressed by WINi in responsive cancer cell lines.

Together, these data reveal that improvements in the potency of WINi have not resulted in substantive changes in their impact on the transcriptome of MLLr cells, and reinforce the concept that inhibition of select RPG expression—and induction of a p53-related transcriptional program—defines the response of the transcriptome to WIN site blockade in this setting.

Impact of WINi on the translatome of MLLr cancer cells

We previously showed that treatment of MV4;11 cells with C6 results in a time-dependent decrease in translation, as measured by bulk labeling of nascent polypeptide chains with O-propargyl-puromycin (OPP) (Aho et al., 2019a). We confirmed this finding with C6, and extended it to C16, showing that while there is no significant effect on protein synthesis capacity after 24 hours of WINi treatment, a progressive decline begins at 48 hours, reaching a ∼40% reduction at the 96 hour treatment point (Figure 2—figure supplement 1A–B). To determine more precisely the effects of WIN site inhibition on translational processes, we performed ribosome profiling [Ribo-Seq; (McGlincy and Ingolia, 2017)] in parallel with the RNA-Seq analyses described above. Spike-in controls were not included. By sequencing ribosome protected fragments (RPF) in 48 hour-treated and control cells, and normalizing to transcript levels from RNA-Seq, we calculated the translation efficiency (TE) of each transcript and used this to determine how C6/C16 influence translation, independent of effects on mRNA abundance.

Impact of WINi on the translatome of MLLr cancer cells.

(A) Volcano plots depicting alterations in translation efficiency (TE) induced by 48 hour treatment of MV4;11 cells with either 2 µM C6 (left) or 100 nM C16 (right) compared to DMSO (n = 2; Red indicates FDR < 0.05 and Log2 FC > 0.25), as determined by Ribo-Seq. (B) Number of mRNAs with significantly (FDR < 0.05 and Log2 FC > 0.25) altered TE levels following treatment of MV4;11 cells with C6 (2 µM) or C16 (100 nM) for 48 hours. See Figure 2—source data 1 for complete output of Ribo-Seq analysis. (C) Overlap of mRNAs with significantly decreased TE in response to C6- or C16-treatment. (D) Translation efficiencies (TE) of mRNAs in DMSO-treated MV4;11 cells plotted against translation efficiencies of mRNAs in cells treated with either C6 (left) or C16 (right). Red indicates mRNAs with significantly altered translation efficiencies following inhibitor treatment (FDR < 0.05 and Log2 FC > 0.25). (E) Numbers of differentially-translated mRNAs (ΔTE) in each quartile of genes (stratified by TE in DMSO) in cells treated with C6 (left) or C16 (right). (F) Enrichment analysis of common mRNAs suppressed by C6/C16 at the mRNA (blue) and translational (red; TE) level in MV4;11 cells. Hallmark.MSigDB pathways are shown. The x-axis indicates the number of suppressed genes in each category; the italic numbers are the corresponding FDR. See Figure 2—source data 2 for the full Hallmark.MSigDB analysis, as well as for Reactome and KEGG pathways. (G) Enrichment analysis of mRNAs suppressed translationally by C6/C16 but with no significant changes in mRNA levels. Gene Ontology (GO) Biological Process (BP) and Molecular Function (MF) categories are shown, as well as KEGG pathways. The x-axis displays -Log10 FDR; the number of mRNAs is shown in italics in each bar. See Figure 2—source data 3 for extended enrichment analyses, broken down by TE and mRNA direction changes. (H) TE changes in WDR5-bound (left) and non-bound (right) RPGs elicited by C6 (top) or C16 (bottom).

In these experiments, RPFs have the characteristic length of ribosome-protected mRNA fragments (28–32 nucleotides; Figure 2—figure supplement 2A), are enriched in coding sequences (Figure 2—figure supplement 2B), and map to the expected reading frame Figure 2—figure supplement 2C), all of which indicate successful profiling. In contrast to RNA-Seq, where we see equal numbers of transcript increases and decreases in response to C6/C16, the overwhelming effect of these compounds on translation efficiency is inhibitory (Figure 2A). Of the ∼10,000 transcripts profiled, between ∼4,500 (C16) and ∼5,900 (C6) transcripts show decreased TE, compared to less than 10 transcripts with increased translation (Figure 2B and Figure 2—source data 1). As we observed in the RNA-Seq, changes in TE are generally less than two-fold (Figure 2A) and there is extensive overlap between the two inhibitors, with ∼90% of the transcripts decreased in TE by C16 also decreased by C6 (Figure 2C). In general, C6/C16 reduce translation of mRNAs in a manner independent of basal translation efficiencies (Figure 2D), although if we bin transcripts according to basal TE we observe that the number of highly-translated transcripts (fourth quartile) impacted by C6/C16 is greater than for those transcripts with lower basal TE (Figure 2E). Within these quartiles, however, the magnitude of reduction in TE is equivalent (Figure 2—figure supplement 2D). Interestingly, mRNAs carrying better matches to the 5’TOP motif—which links translation to mTORC1 signaling (Philippe et al., 2020)—show less decrease in TE compared to those with poorer matches (Figure 2—figure supplement 2E), suggesting that mTORC1-regulated mRNAs may be spared from the full translational effects of WIN site inhibition.

Interrogating specific changes in TE induced by C6/C16, we see that the biological categories of transcripts with decreased TE echo many of those observed with decreased mRNA levels, but include more genes in each category. For example, manually curating each list for transcripts encoding the ∼60 validated substrates of the protein arginine methyltransferase PRMT5 (Radzisheuskaya et al., 2019) reveals that 42 are translationally suppressed by C6/C16, compared to just 10 that are suppressed at the mRNA level (Figure 2—figure supplement 2F). Probing for Hallmark categories in the Human Molecular Signatures Database [MSigDB; (Liberzon et al., 2015)] uncovers the extent of this phenomenon (Figure 2—source data 2), with categories linked to MYC, E2F, mTORC1 signaling, and the G2M checkpoint (Figure 2F) all represented by more genes in the ribosome profiling than in the RNA-Seq experiments. Within these categories, a majority of genes suppressed by C6/C16 at the mRNA level are also further suppressed translationally (Figure 2—figure supplement 2G). This is not a general trend in the response, however, as fewer than half of the total transcripts with reduced mRNA abundance experience this additional translational inhibition (Figure 2—figure supplement 2H). The finding that TE changes induced by C6/C16 extend the biological characteristics of changes in mRNA abundance may indicate a role for impaired translation in contributing to at least some of the mRNA level differences triggered by WINi. Indeed, comparing transcripts with decreased TE but no mRNA decrease with transcripts with decreased TE and mRNA levels reveals the latter are enriched in so-called "optimal codons" (Wu et al., 2019) that normally promote mRNA stability but are linked to mRNA instability when translation is inhibited (Figure 2—figure supplement 2I).

Implicit in the previous discussion, a majority of the translational decreases triggered by C6/C16 occur at transcripts for which there are no significant changes in mRNA abundance (Figure 2—figure supplement 2H). Expectedly, genes with decreased TE but no mRNA level changes are enriched in several of the major Hallmark categories described above (Figure 2—source data 3). But we also observe enrichment in genes connected to the proteasome, spliceosome, mRNA surveillance, and translation (Figure 2G). The latter category includes subunits of the mitochondrial ribosome (Figure 2—figure supplement 3A), translation and ribosome biogenesis factors, and an expanded cohort of transcripts from RPGs (Figure 2H). Indeed, compared to mRNA levels, where C6/C16-induced changes are confined (with the exception of RPL22L1) to a decrease in expression of WDR5-bound RPGs, translational effects are not, and some of the most pronounced TE changes occur at non-WDR5 ribosomal protein target genes. Thus, beyond what we have been able to infer from previous studies, WINi causes a widespread reduction in the ability of MLLr cells to efficiently translate mRNAs connected to almost every aspect of protein synthesis and homeostasis.

Impact of WINi on the ribosome inventory of MLLr cancer cells

Based on the finding that WDR5 controls expression of half the RPGs, we speculated that WINi induce a ribosome subunit imbalance that leads to induction of p53 (Aho et al., 2019a). It is also possible, however, that quality control mechanisms deplete the entire inventory of ribosomal proteins (RP) during prolonged WIN site blockade. To distinguish between these possibilities, we tracked changes in ribosomal protein levels at two timepoints: 24 hours, when there is no overt cellular response to WIN site inhibition, and 72 hours, when cell proliferation begins to be inhibited (Aho et al., 2019a). The abundance of ribosomal proteins allows for the use of label-free quantitative mass spectrometry [LFQMS; (Cox et al., 2014)] in whole cell lysates to feasibly track ribosome protein levels, while also providing insight into other changes in protein levels promoted by WIN site inhibitor. Spike-in controls were not included.

In this analysis, we tracked ∼3,200 proteins at each timepoint (Figure 3A and Figure 3—source data 1), ∼850 of which are significantly altered by C16 treatment. Consistent with the subtle effects of WINi on mRNA abundance and TE, most differences in protein levels triggered by C16 are less than two-fold in magnitude (Figure 3B and 3C). At 24 hours, ∼90% of the proteins that change in response to C16 score as increased, whereas by 72 hours this number drops to ∼60% (Figure 3A). Although the induction of proteins in response to C16 is unexpected for agents that decrease translational capacity, we note that protein synthesis is largely unaffected after 24 hours of C16 treatment (Figure 2—figure supplement 1), and that this phenomenon is unlikely to be an artifact of normalization. Indeed, the distribution of peptide intensities in the LFQMS data is unaffected by normalization (Figure 3—figure supplement 1A), and we see almost as many highly abundant proteins increasing as decreasing with C16 treatment (Figure 3—figure supplement 1B–D), arguing against the idea that large decreases in highly expressed proteins (such as the RPs; Figure 3—figure supplement 1B) are creating the appearance of less abundant proteins being induced. Instead, we suggest that this may be a transient compensatory mechanism, or an early part of the response to WINi. Regardless, most instances of increased protein levels are transient—fewer than one third of these proteins are still induced at day three (Figure 3D)— whereas a majority of the proteins decreased at day one are also decreased at day three. Enrichment analysis (Figure 3—figure supplement 2A and Figure 3—source data 2) reveals that proteins induced at 24 hours are modestly enriched in those connected to exocytosis and leukocyte activation, as well as mTORC1 signaling and MYC. By 72 hours, we see induction of proteins linked to glycolysis and fatty acid metabolism, as well as apoptosis. Additionally, manual curation reveals that the number of induced p53 target proteins increases over time: 13 are induced at 24 hours, compared to 24 at the three day point (Figure 3E). Commensurate with the onset of a functional response to WINi, therefore, is a modest expansion in the apparent impact of p53 on the proteome, as well as the emergence of apoptotic response indicators.

Impact of WINi on the ribosome inventory of MLLr cancer cells.

(A) Lysates from MV4;11 cells treated 24 or 72 hours with either 0.1% DMSO or 250 nM C16 were subjected to liquid chromatography coupled with tandem mass spectrometry and analyzed by label-free quantification (LFQMS). The table shows the number of proteins detected in DMSO and C16 samples and those with significantly altered levels at each time point (n = 4; adj. p-value < 0.05). See Figure 3—source data 1 for complete output of LFQMS analysis. (B) Volcano plot, showing protein level alterations in cells treated with C16 for 24 hours (red indicates adj. p-value < 0.05). The location of RPL22L1 is indicated. (C) As in (B) but for 72 hour treatment with C16. (D) Overlap of proteins significantly increased (top) or decreased (bottom) following 24- or 72-hour C16-treatment. (E) Protein level alterations induced by C16 in consensus p53 target proteins (Fischer, 2017) at the 24 and 72 hour treatment timepoints. Those proteins only altered in abundance at 24 hours are represented as blue dots; proteins only altered at 72 hours are red; proteins altered at both timepoints are grey. (F) As in (E) but for ribosomal proteins. (G) Changes in expression of proteins encoded by WDR5-bound (left) and non-bound (right) RPGs elicited by 24 (top) or 72 (bottom) hour treatment with C16. Note that, due to the magnitude of change, Log2(FC) for RPL22L1 is presented on a separate scale.

Not surprisingly, proteins that are reduced in abundance at 24 hours are significantly enriched in those linked to the ribosome (Figure 3—figure supplement 2B). This enrichment becomes stronger at 72 hours. We also observe, at 72 hours, suppression of proteins linked to MYC and E2F targets, as well as mTORC1 signaling. In terms of ribosome components, this analysis reveals a progressive decline in the ribosomal protein inventory. Going from 24 to 72 hours, there is an increase in the number of impacted ribosomal subunits as well as in the magnitude of their suppression (Figure 3F), and eventually almost all ribosomal subunits are in deficit, regardless of whether or not they are encoded by a WDR5-bound gene (Figure 3G). RPL22L1 is the most strongly suppressed protein at 72 hours (Figure 3C), with its levels reduced by an order of magnitude after three days of C16 treatment. Consistent with the highly coordinated nature of ribosome biogenesis (Dorner et al., 2023), decreases in the abundance of RP are accompanied by a progressive decline in rRNA expression—as revealed by metabolic labeling of RNAs with 2’-azido-2’-cytidine (AzCyd) (Figure 3—figure supplement 3 and Figure 3—source data 3). Together, these experiments reveal that changes in ribosomal protein levels predicted from our transcriptomic studies manifest in reduced expression of ribosome components. Contrary to our earlier idea that WINi promote ribosome subunit imbalance, however, these data support a simpler model in which these inhibitors ultimately induce attrition of the majority of ribosomal proteins—as well as mature rRNAs.

Finally, we asked if the decline in ribosome inventory triggered by C16 is associated with nucleolar stress, as we have shown with C6 (Aho et al., 2019a). We used immunofluorescence to measure the redistribution of nucleophosmin (NPM1) from the nucleolus to the nucleoplasm; a characteristic of this phenomenon (Russo and Russo, 2017). Because inhibition of ribosome biogenesis via some rRNA inhibitors can induce a DNA damage response (Sanij et al., 2020), we also probed for the DNA damage marker ɣ-H2AX. We observed no obvious change in nucleolar morphology with up to 72 hours of C16 treatment (Figure 3—figure supplement 4A). We did, however, see a significant decrease in the nucleolar enrichment of NPM1 at the 72 hour treatment timepoint (Figure 3—figure supplement 4B), indicative of a nucleolar stress response. Notably, we did not observe induction of ɣ-H2AX foci in either the nucleolus or nucleoplasm, detecting it only in cells that were morphologically apoptotic (Figure 3—figure supplement 4A), consistent with studies showing that ɣ-H2AX is induced via the DNA fragmentation that occurs during apoptosis (Rogakou et al., 2000). Based on this analysis, we conclude that activation of nucleolar stress occurs in response to prolonged exposure to C16 and that, compared to some rRNA inhibitors, widespread induction of DNA damage is not a specific consequence of the action of WINi.

A loss of function screen for modulators of the response to WINi

Next, we conducted a two tier loss of function screen to identify genes that modulate the response of MLLr cells to WINi (Figure 4A). Our objective was to compare C6 and C16, and to identify high-confidence hits that are disconnected from cell viability. In Tier 1, we carried out a screen using the GeCKOv.2 sgRNA library (Joung et al., 2017), which targets ∼19,000 genes with six sgRNAs each, as well as ∼1,200 miRNAs (four sgRNAs each). After transducing the library into MV4;11 cells expressing Cas9, we treated for two weeks with 2 μM C6, during which time rapidly growing cells emerged within the transduced population (Figure 4—figure supplement 1A). We harvested genomic DNA, performed next-generation sequencing, and compared sgRNA representation before and after C6 treatment. We then inventoried genes with significant enrichment/depletion in corresponding sgRNAs in the treated population, removed pan-essential genes (Tsherniak et al., 2017), and created a custom library in which non-essential protein-coding "hits" are targeted by four different sgRNAs (Doench et al., 2016). The smaller Tier 2 library was then screened against C6 or C16, this time against a parallel DMSO-treated control population. This two-tiered approach allowed us to efficiently screen two different WINi and to identify hits that are validated (for C6 at least) with up to 10 unique sgRNAs.

A two tier loss of function screen for modulators of the response to WINi.

(A) Two-tier screen design. In the first tier, Cas9-expressing MV4;11 cells were transduced with a genome-wide sgRNA library and treated with 2 µM C6 until a resistant cell population emerged. sgRNA representation in the pre-treatment population was compared to the post-treatment population. In the second tier, cells were transduced with a custom library of distinct sgRNAs targeting non-pan-essential "hits" from the first tier, cultured in the presence of DMSO, C6, or C16, and sgRNA representation in C6/C16-treated cultures compared to that from DMSO-treated cultures. Created with BioRender.com. (B) Volcano plot, showing gene-level changes in sgRNA representation from the first tier (orange indicates FDR < 0.05). Datapoints corresponding to TP53, RPL22, and CDKN2A are indicated. See Figure 4—source data 1 for full output of the Tier 1 screen. (C) Comparison of gene-level changes in sgRNA representation in C6- and C16-treated populations in the second tier screen, each compared to DMSO-treated populations (red indicates FDR < 0.05; black indicates non-targeting control sgRNAs). See Figure 4—source data 2 for full output of the Tier 2 screen. (D) Top: Overlap of genes from the Tier 2 screen with enriched (left) or depleted (right) sgRNAs in C6- and C16-treated MV4;11 populations, compared to the DMSO control. Bottom: Overlap of genes with enriched (left) or depleted (right) sgRNAs in the first versus second tiers of the screen. "Tier 1" contains only those genes targeted in the Tier 2 screen. "Tier 2" contains the intersection of genes with altered sgRNAs in both the C6 and C16 treatments. (E) Ranked heatmap, representing the mean gene-level Log2 fold change (FC) of sgRNAs from the C6 and C16 treatments in the Tier 2 screen, as well as gene enrichment analysis outputs. Note that "Signal transduction by p53 class mediator" is a GO:BP term (orange); "p53" assignments (yellow) were added by manual curation.

Although the first tier did not discriminate between genes that modulate fitness and those that modulate WINi response, several interesting observations emerged. Guide RNAs corresponding to ∼70 genes were enriched and ∼675 were depleted (Figure 4B), most of the latter of which are pan-essential (Figure 4—source data 1). Satisfyingly, TP53 is the most highly enriched gene in the screen (Figure 4B and Figure 4—figure supplement 1B). CDKN2A also scored as highly enriched in the initial screen (Figure 4B)—specifically those sgRNAs targeting p14ARF(Figure 4—figure supplement 1C–1D), an inhibitor of the p53 ubiquitin ligase MDM2 (Sherr, 2001). Further support for the importance of p53 is evidenced by network analysis (Chang and Xia, 2023) of the 27 miRNAs flagged as enriched (Figure 4—source data 1), which display connections to p53 (Figure 4—figure supplement 1E). Finally, we note that the second most highly-enriched gene in the first tier encodes a ribosomal protein: RPL22 (Figure 4B and Figure 4—figure supplement 1F). Because of the strong enrichment of sgRNAs against RPL22 and TP53, we removed both genes from the second tier screen.

The second tier screen (Figure 4—figure supplement 1G and Figure 4—source data 2) revealed that the response of MV4;11 cells to C6 and C16 is very similar, both in terms of the enriched/depleted genes and their rankings (Figure 4C). A majority of genes that modulate the response to C6 similarly modulate the response to C16 (Figure 4D). For most of the depleted genes that appear specific to one WINi, similar depletion is observed with the other WIN site inhibitor, but is generally just over the FDR cutoff (Figure 4—figure supplement 1H). But for C6-specific enriched genes we see that most have high FDR values in the C16 samples, arguing that the earlier generation compound has expanded, off-target, activities. Gene ontology (GO) enrichment analysis of the 57 common genes emerging from the screen revealed enrichment in four overlapping categories connected to p53 signaling, apoptosis, the DNA damage response (DDR), and histone modifications (Figure 4E and Figure 4—figure supplement 1I). The representation of genes connected to p53 and to apoptosis reinforces the importance of p53-mediated cell death to the response of MLLr cells to WINi. We observe, for example, that loss of function of the p53 antagonist and ubiquitin ligase MUL1 (Jung et al., 2011), increases sensitivity to C6/C16, whereas loss of canonical p53 effectors NOXA, BAX, and ZMAT3 is associated with a decrease in response. The DNA damage response category overlaps with that of p53 but is nonetheless distinct and includes genes encoding the ATM and CHK2 kinases (Blackford and Jackson, 2017) and the FOXM1 transcription factor that activates DDR gene expression networks (Zona et al., 2014). This category also includes two depleted genes, encoding FBXL5—which antagonizes ATM signaling (Chen et al., 2014)—and DYRK1A—a kinase involved in the DDR (Laham et al., 2021), DREAM complex activation (Litovchick et al., 2011), and RPG transcription (Di Vona et al., 2015). The involvement of this category of enriched genes is intriguing, given the lack of ɣ-H2AX accumulation in non-apoptotic cells (Figure 3—figure supplement 4), and warrants further investigation in the future.

A majority of genes in the histone modification category, when disrupted, blunt the response to both WINi (Figure 4E). These genes include those encoding the H3K27 demethylase KDM6A (Lan et al., 2007), the MLL/SET protein KMT2D [MLL2; (Shinsky et al., 2015)], and ARID4A—a component of the mSin3/HDAC1 co-repressor complex (Lai et al., 2001). The most conspicuous sensitizing gene in this group is BRD3, a member of the BET family of proteins that includes BRD2 and BRD4 (Eischer et al., 2023). Interestingly, although BRD4 was not included in the second screen tier as it is pan essential, BRD2 was not included because it was not significantly enriched/depleted in the first tier (Figure 4—source data 1), revealing that the actions of BRD3 in modulating response to WINi are not shared with all family members. Further supporting the importance of BRD3 to the response, we note that SPOP, which targets BET family proteins for proteasomal destruction (Janouskova et al., 2017), is one of the most significantly enriched hits from the screen (Figure 4E).

Collectively, these findings demonstrate functional involvement of the ribosomal protein RPL22 in the response to WINi, and confirm the importance of the p53 network to robust inhibition of MLLr cell growth by these agents. These findings also identify a number of candidate predictive biomarkers of response that can be further interrogated for their value as patient-selection criteria.

Identification of agents that synergize with WINi in MLLr cells

Given the ways in which resistance to WINi can arise (Figure 4), the most efficacious application of these agents will likely be in combination with other therapies. We therefore asked whether C16 synergizes with 11 approved or targeted agents. Several of the agents were chosen based on the results of our screen. Harmine—an inhibitor of the DYRK1A kinase (Gockler et al., 2009)—and the BET family inhibitor mivebresib (McDaniel et al., 2017) each target the product of sensitizing genes, whereas venetoclax inhibits BCL-2 (Souers et al., 2013)—an inhibitor of BAX, which scored as a resistance gene. We also tested agents connected to the DDR [etoposide, olaparib, and the ATR inhibitor VE821; (Charrier et al., 2011)], protein synthesis and homeostasis (alvespimycin and rapamycin), and p53 (nutlin-3a). Due the the enrichment of PRMT5 substrates in our translational profiling (Figure 2—figure supplement 2F), we queried the PRMT5 inhibitor pemrametostat (Chan-Penebre et al., 2015). And because DOT1L inhibitors suppress not only classic MLL fusion target genes (Bernt and Armstrong, 2011, Daigle et al., 2011) but also RPGs (Lenard et al., 2020), we tested for synergy with the DOT1L inhibitor pinometostat (Daigle et al., 2011). We treated MV4;11 cells with a dose matrix spanning 49 unique dose combinations and quantified synergy δ-scores using the Zero Interaction Potency (ZIP) model (Yadav et al., 2015) (Figure 5A-B, Figure 5—figure supplement 1A and 1B, and Figure 5—source data 1).

Identification of agents that synergize with WINi in MLLr cells.

(A) Peak synergy (> 0) and antagonism (< 0) ZIP Delta (δ) scores from synergy assays in which MV4;11 cells were treated for three days with 49 unique dose combinations of C16 and the indicated compound of interest (n = 4). See Figure 5—source data 1 for numerical ZIP Delta analysis output. (B) Heatmaps of MV4;11 cell growth inhibition at each dose of C16 and the indicated six compounds. The remaining five combinations tested are shown in Figure 5—figure supplement 1. (C) As in (A) but for MOLM13 cells. See Figure 5—source data 1 for numerical ZIP Delta analysis output. (D) As in (B) but for MOLM13 cells. The remaining five combinations tested are shown in Figure 5—figure supplement 2. (E) Number of genes with significantly (FDR < 0.05) altered transcript levels following treatment of MV4;11 cells with C16 (100 nM), mivebresib (Mibv; 2.5 nM), or the combination for 48 hours, as determined by RNA-Seq (n = 3). See Figure 5—source data 2 for complete output of RNA-Seq analysis. (F) UpSet plot, showing the overlap of genes suppressed (left) or induced (right) in response to C16, mivebresib, or the combination. (G) UpSet plot, showing the breakdown of Reactome "Translation" pathway genes suppressed in response to C16, mivebresib, or the combination. (H) Enrichment of Reactome Pathways in genes with increased transcripts following treatment of MV4;11 cells with C16, mivebresib, or the combination. See Figure 5—source data 3 for complete output of enrichment analyses.

In MV4;11 cells, we observe synergy with mivebresib, pemrametostat, pinometostat, etoposide, harmine, and venetoclax. Within this group are three agents selected based on sensitizing targets from the CRISPR screen, providing additional support for the role of DRYK1A, BCL-2/BAX, and BRD3 in the responsiveness to WINi. Of the three agents connected to the DDR, only etoposide displays significant synergy. Agents that target protein synthesis and homeostasis yield mixed results—we observe potent antagonism with the mTOR inhibitor (Raught et al., 2001) rapamycin (peak δ-score -16), while the HSP90 inhibitor alvespimycin (Schnur et al., 1995) is either antagonistic or synergistic, depending on dose (Figure 5A). Finally, we note that—of the agents displaying synergy—four are particularly strong (peak δ-scores >10) and observed at agent doses consistent with on-target activity (Figure 5—figure supplement 1B), suggesting that mivebresib, pemrametostat, pinometostat, and venetoclax should be prioritized for in vivo testing. Focusing on these agents is further justified by our finding that all four are synergistic with C16 in MOLM13 cells (Figure 5C–D and Figure 5—figure supplement 2A and 2B).

To understand how combination with another agent impacts the response to WINi, we transcriptionally profiled MV4;11 cells treated for forty-eight hours with C16 and mivebresib, either as single agents or in combination, at concentrations that yield peak synergy between them (100 nM C16 and 2.5 nM mivebresib). Spike-in controls were not included. By RNA-Seq (Figure 5—source data 2), it is clear that the functional synergy between C16 and mivebresib is apparent at the transcript level, with more than 6,200 gene expression changes in the combination treatment, compared to less than 1,800 for C16 and 2,700 for mivebresib (Figure 5E). Notable are the very distinct transcriptional profiles induced by each agent alone, with fewer than 200 shared gene expression changes in each direction (Figure 5F). The impact on RPG expression of both agents is additive (Figure 5—figure supplement 3A), but in general we find that the combination of C16 and mivebresib dysregulates similar categories of genes for each agent alone, but with substantially more genes in each category (Figure 5—source data 3). This is clear for genes linked to translation (Figure 5G), to p53 (Figure 5—figure supplement 3B), and to the induction of apoptosis (Figure 5H). Thus, although further investigation is needed, this analysis is consistent with the idea that synergy between C16 and mivebresib results from alterations in the expression of distinct but complementary sets of genes that ultimately conspire to augment induction of p53.

WINi inactivate MDM4 in an RPL22-dependent manner

Despite the importance of p53 in the response to WINi, WIN site inhibitors cause only a slight increase, if any, in p53 levels [Figure 6—figure supplement 1A, Figure 6—source data 1 and (Aho et al., 2019a)]. Interestingly, inactivation or loss of RPL22 in cancer is associated with increased expression of RPL22L1 and inclusion of exon 6 in MDM4 (Ghandi et al., 2019), an event that promotes MDM4 expression by preventing formation of a "short" MDM4 mRNA isoform (MDM4s) that is destroyed by nonsense-mediated decay (Rallapalli et al., 1999). MDM4 is intriguing because it can suppress p53 without altering its stability (Francoz et al., 2006). It is also intriguing because skipping of exon 6 in the MDM4 mRNA is stimulated by ZMAT3 (Bieging-Rolett et al., 2020, Muys et al., 2021) and antagonized by RPL22L1 (Larionova et al., 2022)—two genes that are oppositely regulated by WINi. We therefore asked if WINi induce changes in the levels of mRNA splice isoforms and if this includes MDM4.

WINi inactivate MDM4 in an RPL22-dependent manner.

(A) Differential alternative splicing events affected by C6/C16 treatment of MV4;11 cells were quantified by rMATS. The types of alternative splicing events are cartooned at left, and the number of significantly different events (> 5% Δψ; FDR < 0.05) common to C6/C16 depicted in the graph. See Figure 6—source data 2 for output of rMATS analysis. (B) Sashimi plot quantifying read junctions that span exons 5–7 of MDM4 in MV4;11 cells treated with DMSO (green) or C16 (blue). Numbers in the arcs display junction depth. The location of exons 5, 6, and 7 is depicted at the bottom; skipped exon 6 is highlighted in orange. (C) Viabilities of control (non-targeting: NT) and RPL22 knock out (KO) MV4;11, MOLM13, and K562 cells treated with a serial dilution range of C16 for 72 hours, relative to viability of DMSO-treated cells (n = 3; Mean ±SEM). (D) Western blot analysis of p53 levels in control (NT) and RPL22 knockout (KO) MV4;11 and MOLM13 cells treated with either 0.1% DMSO or C16 (MV4;11, 200 nM; MOLM13, 400 nM) for 72 hours. α-Actinin is loading control. Representative images from three biological replicates shown. Raw unprocessed gel images are presented in Figure 6—source data 5. (E) Heatmap, showing significant changes in the expression of consensus p53 target genes (Fischer, 2017) between the indicated pairwise comparisons of RNA-Seq datasets. Note that only consensus p53 target genes altered in expression by C16 in control (NT) cells are represented. (F) Sashimi plot quantifying read junctions that span exons 5–7 of MDM4 in RPL22KO MV4;11 cells treated with DMSO or C16. Numbers in the arcs display junction depth. The location of exons 5, 6, and 7 is depicted at the bottom; skipped exon 6 is highlighted in orange. Corresponding NT images are presented alongside RPL22KO images in Figure 6—figure supplement 3B). (G) Western blots, comparing the effects of 72 hours of DMSO (DM) or C16 treatment (MV4;11, 200 nM; MOLM13, 400 nM) of control (NT) or RPL22 knockout (KO) MV4;11 (left) or MOLM13 (right) cells on levels of MDM4, p21, RPL22L1, RPL22, and GAPDH (loading control). Representative images from three biological replicates are shown. Raw unprocessed gel images are presented in Figure 6—source data 9.

RNA-Seq data (Figure 1) were interrogated for alternative splicing events (Shen et al., 2014). At an FDR < 0.05 and a threshold of ≥ 5% change in exon inclusion (Δψ), C6 and C16 each result in changes in ∼1,000 differentially-spliced mRNAs (Figure 6—figure supplement 1B), ∼250 of which are shared between the two inhibitors (Figure 6A). Many of these changes reflect events with low read counts or at minor splice sites (Figure 6—source data 2), representative examples of which are presented in Figure 6–figure supplement 1C–D. That said, WINi clearly promote accumulation of MDM4 transcripts in which exon 6 is skipped (Figure 6B). We also observe splicing changes at RPL22L1 itself (Figure 6–figure supplement 1E), where WINi leads to the depletion of transcripts in which exon 2 is spliced to a distal 3’ acceptor site in exon 3. This splicing event encodes the RPL22L1a isoform that is incorporated into ribosomes (Larionova et al., 2022). Splicing to the proximal 3’ acceptor site, which generates a non-ribosomal RPL22L1b isoform that modulates splicing, is insensitive to WINi. We confirmed the impact of C16 on MDM4 and RPL22L1 splice isoforms by semi-quantitative RT-PCR (Figure 6—figure supplement 1F and Figure 6—source data 3) and quantitative RT-PCR (Figure 6—figure supplement 1G). Based on these observations, we conclude that treatment of MLLr cells with WINi promotes the selective loss of transcripts encoding RPL22L1a and MDM4.

The association of RPL22 loss with increased expression of RPL22L1 and inclusion of exon 6 in MDM4 (Ghandi et al., 2019) prompted us to ask how RPL22 contributes to the response of MLLr cells to WINi. Knockout (KO) of RPL22 (Figure 6—figure supplement 2A and Figure 6—source data 4) decreases the sensitivity of MV4;11 and MOLM13 cells to C16 by three- to five-fold compared to non-targeted (NT) control cells (Figure 6C and Figure 6—figure supplement 2B), as well as attenuating the modest induction of p53 protein observed in the MV4;11 line (Figure 6D and Figure 6—source data 5). The response of relatively insensitive (p53-null) K562 cells, in contrast, is unaffected by RPL22 disruption (Figure 6C). RNA-Seq analysis, performed without spike-in controls, (Figure 6—figure supplement 2C–D and Figure 6—source data 6) reveals that disruption of RPL22 does not impact the effect of WINi on WDR5-bound RPGs (Figure 6—figure supplement 2E), but it does block the effects of C16 on expression of RPL22L1 and ZMAT3 (Figure 6—figure supplement 2D), as well as tempering its ability to suppress genes connected to the cell cycle, mTORC1 signaling, and MYC (Figure 6—figure supplement 2F and Figure 6—source data 7). Notably, RPL22 loss also impairs induction of genes involved in p53 signaling (Figure 6E and Figure 6—figure supplement 2G). We also observe that mitochondrial RPGs are induced by WINi uniquely in RPL22-null cells (Figure 6—figure supplement 2H). We conclude that RPL22 is needed for a majority of the characteristic responses of MLLr cells to WINi, including activation of p53.

Finally, we asked if RPL22 knockout alters patterns of alternative splicing induced by WINi. Thousands of differences were detected in splice isoforms between the various pairwise comparisons (Figure 6—source data 8). In general, RPL22KO cells show fewer C16-induced changes in alternative splicing patterns than NT cells (Figure 6—figure supplement 3A). As we observed above, a majority of the changes reflect events with low read counts or at minor splice sites, with two notable exceptions: MDM4 and RPL22L1. In the absence of WINi, disruption of RPL22 promotes exon 6 retention in MDM4 (Figure 6—figure supplement 3B) and induces expression of the mRNA splice isoform encoding RPL22La (Figure 6—figure supplement 3C). In the presence of C16, disruption of RPL22 mitigates skipping of exon 6 in MDM4 (Figure 6F) and largely blocks suppression of the RPL22L1a-encoding mRNA isoform (Figure 6—figure supplement 3C). Importantly, these changes in mRNA isoforms manifest at the protein level (Figure 6G and Figure 6—source data 9) as we observe that both MDM4 and RPL22L1 protein expression is suppressed by C16, that RPL22L1 is induced by RPL22 disruption, and that loss of RPL22 blocks a majority of the ability of C16 to reduce expression of the MDM4 and RPL22L1 proteins. We also note that loss of RPL22 blocks induction of the p53 target, p21. Taken together, these data demonstrate that RPL22 loss can cause the induction of RPL22L1 and inclusion of exon 6 in MDM4 observed in cancer (Ghandi et al., 2019), and reveal that splicing-driven suppression of MDM4 is an important mechanism through which WINi activate p53 in MLLr cells.

Discussion

Here, we describe an integrated multiomic approach to characterize the mechanism of action of WDR5 WIN site inhibitors in MLL-rearranged cancer cells. By combining transcriptional, translational, and proteomic profiling with genome-wide loss of function screens, we demonstrate the broad impact of WINi on the ribosomal protein complement and translational capacity of MLLr cells, reveal the importance of multiple arms of the p53 response pathway in cellular inhibition, and uncover a role for alternative splicing of MDM4 in activating p53 in this setting. Collectively, these data cast WIN site inhibitors as a novel ribosome-directed anti-cancer therapy and provide insight into patient selection criteria, mechanisms of resistance, and strategies to improve WINi efficacy in the clinic.

The primary targets of WINi, we propose, are the cohort of ∼40 RPGs that are bound by WDR5 in all cell types examined (Aho et al., 2019a, Bryan et al., 2020, Florian et al., 2022). These RPGs are rapidly transcriptionally suppressed in response to WIN site blockade (Bryan et al., 2020, Florian et al., 2022) ( Aho et al., 2019a) or WDR5 degradation (Siladi et al., 2022), and show sustained suppression at the mRNA level. These same RPGs are suppressed by disruption of the MYC–WDR5 interaction (Thomas et al., 2019), implying that the function of WDR5 at these genes is to recruit MYC, a prominent target of MLL-fusion oncoproteins (Ahmadi et al., 2021). Previously, we posited that suppression of half the RPGs would lead to a ribosomal subunit imbalance (Aho et al., 2019a), but our analysis shows that WINi deplete the entire ribosomal inventory. Part of this depletion is driven by decreased translation of ribosomal mRNAs, although it is also possible that ubiquitin-mediated quality control mechanisms or ribophagy (Zhao et al., 2022) degrade ribosomal proteins under these conditions.

The effects of WINi on ribosomal protein levels are extensive in terms of the number of impacted subunits, but not the magnitude of change, which caps at around a 30% decrease by 72 hours. This cap could be set by the maximal contribution of MYC–WDR5 to the expression of target RPGs, which is two-fold as determined by genetic disruption of the MYC–WDR5 interaction (Thomas et al., 2019), degradation of WDR5 (Siladi et al., 2022), or WIN site blockade. Although modest, it should be noted that a 30% decrease in ribosomal protein levels corresponds to a loss of up to three million ribosomes per cell (Shore and Albert, 2022), and in that light it is not surprising that WINi result in reduced translational efficiencies of about half of all actively-translated mRNAs. Unlike perturbations such as ER stress (Advani and Ivanov, 2019), WINi does not promote preferential translation of stress-responsive mRNAs, although there is some specificity in terms of the translational consequences. Messenger RNAs carrying 5’TOP motifs, for example, are spared from the full translational impact of C6/C16, and we see distinct biological clustering of mRNAs with decreased translation efficiency. Whether these patterns are intrinsic to WINi, or a general response of MV4;11 cells to translational stress, remains to be determined. Overall, we conclude that WINi do not trigger extensive translational reprogramming, but rather act to induce a widespread yet restrained translational choke.

In addition to ribosomal protein attrition, we also find that WINi trigger a progressive decline in rRNA expression, which we suggest is a secondary effect resulting from the tight coordination between RPG and rRNA transcription (Dorner et al., 2023). And, as previously reported (Aho et al., 2019a), we confirm that they promote a shift in the relative abundance of NPM1 in the nucleoplasm versus the nucleolus; indicative of a nucleolar stress response. Given that we recovered multiple DDR components in our two tier CRISPR screen, and that agents such as the RNA polymerase I inhibitor CX-5461 induce DNA damage (Quin et al., 2016, El Hassouni et al., 2019), we had expected to see some evidence of ɣ-H2A.X induction in the nucleolus or the nucleoplasm in response to WINi. Yet we only observe ɣ-H2A.X induction in apoptotic cells. The difference in this aspect of the response to what are overtly two different ways to inhibit protein synthesis in cancer cells—RPG versus rRNA inhibition—may relate to the different primary mechanism of action of these agents or the magnitude of their effects, which are more subtle with WINi. Alternatively, because not all rRNA inhibitors induce DNA damage [e.g., (Peltonen et al., 2014)], induction of DNA damage may be specific to CX-5461. Further investigation, especially into the significance of DDR components in the response to WINi, is warranted.

Activation of p53 is, however, a major point of convergence of WDR5 and rRNA inhibitors. Not only do we detect activation of p53 target genes in response to WIN site blockade, but we observe synergistic activation of p53 target genes with the BET bromodomain inhibitor mivebresib and suppression of p53 targets upon deletion of RPL22; both of which correlate with cellular sensitivity. We also recover multiple components of the p53 signaling pathway as "resistance" genes in our CRISPR screens, including p53 itself and the splicing factor ZMAT3. Inhibition of rRNA synthesis is thought to activate p53 by generating excess ribosomal proteins that bind to and inactivate MDM2 (Pfister, 2019). It is likely that inhibition of MDM2 contributes to p53 activation in response to C6/C16, as we see a modest increase in p53 levels in MV4;11 cells exposed to WINi, and find that loss of the MDM2 inhibitor p14ARF renders MV4;11 cells less sensitive to C6/C16. Here, however, we also find that there is a second route of p53 activation at work, in which WINi promote accumulation of alternatively spliced isoforms of MDM4 mRNA in which exon 6 is skipped, driving down MDM4 protein levels. Because MDM4 inhibits p53 via proteolysis-independent mechanisms (Francoz et al., 2006), these findings explain how WIN site inhibitors can induce a robust p53 target gene signature in the absence of frank induction of p53 protein. They also point to a dominant role of the MDM4–p53 axis in mediating the response of MLLr cells to WIN site blockade. Whether rRNA inhibition triggers p53 activation by a similar mechanism has, to our knowledge, yet to be reported.

In considering the mechanism through which WINi inactivate MDM4, it is possible that the balance of RPL22L1 and ZMAT3, which are oppositely regulated in response to WIN site inhibition, governs the extent of MDM4 exon 6 inclusion. ZMAT3 is induced in response to p53 activation and promotes skipping of exon 6 in MDM4 (Bieging-Rolett et al., 2020). RPL22L1, in contrast, which is potently suppressed by WIN site inhibitor, promotes exon 6 inclusion (Larionova et al., 2022). Any process that tips the balance in favor of ZMAT3, therefore, would be expected to inhibit MDM4, activate p53, and initiate a feed-forward mechanism that drives ZMAT3 expression, fortifies p53 induction, and commits cells to an apoptotic outcome. Induction of ZMAT3 alone should be sufficient to trigger this circuit, but the conspicuous suppression of RPL22L1 by WIN site inhibitor suggests that this may also contribute to the response. Paradoxically, the isoform of RPL22L1 that is suppressed by WIN site blockade, RPL22L1a, is linked in glioblastoma cells to ribosome function, not splicing (Larionova et al., 2022), while the splicing-relevant RPL22L1b isoform is resistant to WINi. One possibility is that, despite its recurrence and prominence, suppression of RPL22L1a by C6/C16 does not contribute to MDM4 suppression, and induction of ZMAT3 is the critical driving factor. Alternatively, RPL22L1a may indeed act to control splicing in MLLr cells, in contrast to what has been reported in glioblastoma. Further investigation is required.

The RPL22–RPL22L1–MDM4 nexus we encountered has been inferred by genome-wide studies of the Cancer Cell Line Encyclopedia [CCLE; (Ghandi et al., 2019)], and our work here demonstrates that loss of RPL22 causes induction of RPL22L1 and promotes inclusion of exon 6 in MDM4. We have no evidence that RPL22 itself plays a role in the response to WINi in otherwise unperturbed MLLr cells. Rather, we suggest that its recovery as a resistance gene in our CRISPR screen is tied to its ability to suppress RPL22L1 expression (O’Leary et al., 2013), and the ectopic effect of its deletion on preventing WINi-induced RPL22L1a decline. That said, RPL22 status is likely to be highly relevant in terms of patient selection criteria. RPL22 is frequently inactivated or deleted in primary cancer samples (Kandoth et al., 2013, Goudarzi and Lindstrom, 2016, Ghandi et al., 2019), as well as 7% of lines in the CCLE (Cao et al., 2017). Unlike other RPGs, mutation or deletion of RPL22 is not associated with mutational inactivation of p53, and indeed there is a strong tendency for wild-type p53 to be retained in RPL22 mutant/deletion lines (Cao et al., 2017, Ghandi et al., 2019). In practical terms, therefore, cancers that retain wild-type p53 but otherwise are mutated/deleted for RPL22, or overexpress RPL22L1, would not be expected to robustly respond to WIN site inhibitor.

Although p53 is important for the action of WINi in MLLr cells, there are likely other stress response mechanisms that mediate cellular inhibition by these agents. A number of p53-independent nucleolar stress responses have been identified, but these remain mechanistically opaque compared to p53-dependent responses (Boglev et al., 2013, Pfister et al., 2015, Jayaraman et al., 2017). Conversely, we might also expect cells to be able to mount protective responses to WIN site blockade. In this regard, it is curious that four resistance genes identified in our CRISPR screen—UBA6, BIRC6, KCMF1, and UBR4—encode members of a newly-identified BIRC6 ubiquitin-ligase complex (Cervia et al., 2023), the function of which is to prevent aberrant activation of the integrated stress response (ISR). The ISR is a central regulator of protein homeostasis (Costa-Mattioli and Walter, 2020) that drives protective translational reprogramming in response to multiple cellular stresses. There is no indication that the ISR is activated by WINi; indeed the master regulator of ISR, ATF4, is suppressed by C6/C16 at the mRNA and translational levels. But the finding that loss of all four members of the BIRC6 complex blunts the response to WINi implies that ISR activation can be a mechanism through which cells evade the full impact of these agents.

As with most monotherapies, future single agent WINi treatment paradigms are likely to encounter resistance, either by activation of protective responses such as those proposed above, or by mutations in one or more of the resistance genes recovered in our CRISPR screen. Identification of agents that can be used in combination with WINi to increase cancer cell inhibition is thus crucial. Our relatively limited synergy screening identified a number of combinations that should be prioritized for in vivo testing. We found that WINi act synergistically with the BCL-2 inhibitor venetoclax (Souers et al., 2013). This is rationalized by our recovery of BAX as a resistance gene, and is noteworthy because venetoclax is an approved therapy for several blood-borne cancers. We also identified notable synergies with experimental agents targeting BET bromodomain family members, DOT1L, and PRMT5. The combination with mivebresib is rationalized based on identification of BRD3 as a sensitizing gene, and likely results from the ability of C16 and mivebresib to inhibit distinct sets of genes connected to translation, the impact of which is to enhance p53 induction. BET bromodomain inhibitors have struggled somewhat in clinical trials due to dose-limiting toxicities (Shorstova et al., 2021), but their combination with WINi could form the basis of a more effective therapy with less side effects. Moreover, given the mechanism underlying synergy between C16 and BET inhibitors, we would expect this combination to be effective in other wild-type p53 cancer settings where WINi are active, such as neuroblastoma (Bryan et al., 2020) and rhabdoid tumors (Florian et al., 2022). Expanded synergy screening is needed to identify and understand the full spectrum of combination approaches that could be used to ultimately enhance and extend the clinical utility of WIN site inhibitors.

Supporting information